### 2015-02-07 common/lstlang-bomil.tex
Initial commit

\lstdefinelanguage{BOMIL}{%
  morekeywords={%
    addr, alloc, and, any, apply, bool, byte, case, ccall, cont, define, do, double,%
    end, enum, extern, false, float, fun, host_vproc, if, in, inline, int, let, long, module,%
    noreturn, promote, pure, return, short, tag, throw, true, typedef, unwrap, use_rw,%
    vec128, void, vproc, vpload, vpstore, wrap,%
  },%
  keywordsprefix=@,
  % otherkeywords={(,),[,],\{,\},\,,:,...,_,|,=,=>,->,\#,:>},
  sensitive,%
  morecomment=[n]{(*}{*)},%
  morestring=[d]",%
}[keywords,comments,strings]%

### 2015-02-07 common/lstlang-sml.tex
Initial commit

\lstdefinelanguage{SML}{%
  morekeywords={%
    abstype, and, andalso, as, case, datatype, do, else, end, eqtype, exception,%
    fn, fun, functor, handle, if, in, include, infix, infixr, let, local, nonfix,%
    of, op, open, orelse, raise, rec, sharing, sig, signature, struct, structure,%
    then, type, val, where, while, with, withtype,%
  },%
  % otherkeywords={(,),[,],\{,\},\,,:,...,_,|,=,=>,->,\#,:>},
  sensitive,%
  morecomment=[n]{(*}{*)},%
  morestring=[d]",%
}[keywords,comments,strings]%

### 2015-02-07 icfp15-servo/abstract.tex
Initial commit

%!TEX root = paper.tex
%
\begin{abstract}
Inlining is an optimization that replaces a call to a function with that function's 
body.
This optimization not only reduces the overhead of a function call, but can 
expose additional optimization opportunities to the compiler, such as removing
redundant operations or unused conditional branches.
Another optimization, copy propagation, replaces a redundant copy of a still-live 
variable with the original.
Copy propagation can reduce the total number of live variables, reducing register
pressure and memory usage, and possibly
eliminating redundant memory-to-memory copies.
In practice, both of these optimizations are implemented in nearly every
modern compiler.

These two optimizations are practical to implement and effective in first-order 
languages, but in languages with lexically-scoped first-class functions (aka, closures), these optimizations 
are not available to code programmed in a higher-order style.
With higher-order functions, the analysis challenge has been that the 
environment at the call site must be the same as at the closure capture location, 
up to the free variables, or the meaning of the program may change.
Olin Shivers' 1991 dissertation called this family of optimizations 
\emph{\superbeta} and he proposed one analysis technique, called \emph{reflow},
to support these optimizations.
Unfortunately, reflow has proven too expensive to implement in practice.
Because these higher-order optimizations are not available in functional-language
compilers, programmers studiously avoid uses of 
higher-order values that cannot be optimized (particularly in compiler
benchmarks).

This paper provides the first practical and effective technique for 
\superbeta (higher-order) inlining and copy propagation, which
we call unchanged variable analysis.
We show that this technique is practical by implementing it in the 
context of a real compiler for an ML-family language and showing 
that the required analyses have costs below 3\% of the total 
compilation time.
This technique's effectiveness is shown through a set of benchmarks
and example programs, where this analysis exposes additional potential
 optimization sites.
\end{abstract}

### 2015-02-07 icfp15-servo/browsers.tex
Initial commit

%!TEX root = paper.tex

\section{Browsers}
\label{sec:browsers}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End:

### 2015-02-07 icfp15-servo/cfa.tex
Initial commit

%!TEX root = paper.tex
%
\section{Control-Flow Analysis}
\label{sec:cfa}
This section provides a brief background on control-flow analysis (CFA) along with an overview 
of our specific implementation techniques to achieve both acceptable scalability and precision. 
A more general introduction to control-flow analysis, in particular the 0CFA
style that we use, is available in the book by Nielson \etal{}~\cite{principles-prog-analysis}.
For a detailed comparison of modern approaches to control-flow analysis, see Midtgaard's 
comprehensive survey~\cite{midtgaard-cfa-survey}.

In brief, while many others have implemented control-flow analysis in their
compilers~\cite{serrano:cfa-paradigm,mlton-cfa,sub-zero-cfa}, our analysis is
novel in its tracking of a wider range of values --- including boolean values and 
tuples --- and its lattice coarsening to balance performance and precision.

\subsection{Overview}
A control-flow analysis computes a finite map from all of the variables in a program 
to a conservative abstraction of the values that they can take on during the
execution of the code.
That is, it computes a finite map 
\begin{displaymath}
 \mathcal{V} : \mathtt{VarID} \finmap \mathtt{value}
\end{displaymath}
where the $\mathtt{value}$ type is defined as a recursive datatype similar to
that shown in \figref{fig:values}.
The special $\top$ (\lstinline{TOP}) and $\bot$ (\lstinline{BOT}) elements 
indicate either all possible values or no known values, respectively. 
A \lstinline{TUPLE} value handles both the cases of tuples and ML datatype
representations, which by this point in the compiler have been desugared into
either raw values or tagged tuples.
The \lstinline{LAMBDAS} value is used for a set of variable identifiers, all of
which are guaranteed to be function identifiers.
The \lstinline{BOOL} value tracks the flow of literal boolean values through
the program.
\begin{figure}[t]
\begin{lstlisting}[basicstyle=\ttfamily\scriptsize\color{\cdColor}]
  datatype value
    = TOP
    | TUPLE of value list
    | LAMBDAS of CPS.Var.Set.set
    | BOOL of bool
    | BOT
\end{lstlisting}
\caption{
  Abstract values.
}
\label{fig:values} 
\end{figure}%

As an example, consider the following code:
\begin{lstlisting}[basicstyle=\ttfamily\scriptsize\color{\cdColor}]
let fun double (x) = x+x
    and apply (f, n) = f(n)
in
    apply (double, 2)
end
\end{lstlisting}%
After running CFA on this example, we have
\begin{displaymath}
  \begin{array}{rcl}
    \mathcal{V}(\texttt{f}) & = & \texttt{LAMBDAS(\SET{double})} \\
    \mathcal{V}(\texttt{n}) & = & \top \\
    \mathcal{V}(\texttt{x}) & = & \top
  \end{array}
\end{displaymath}%
These results indicate that the variable \lstinline{f} must be bound to the 
function \lstinline{double}.
This example and the \lstinline{value} representation in \figref{fig:values} do not
track numeric values, which is why \lstinline{n} and \lstinline{x} are mapped to $\top$.
We are planning to track a richer set of values, including datatype-specific values, in the future
in order to enable optimizations beyond the ones discussed in this paper.

\subsection{Implementation}
Our CFA implementation is straightforward and similar in spirit to Serrano's~\cite{serrano:cfa-paradigm}.
We start with an empty map and walk over the intermediate representation of the 
program.
At each expression, we update $\mathcal{V}$ by merging the value-flow information
until we reach a fixed point where the map no longer changes.
The most interesting difference from Serrano's implementation is that
we use our tracked boolean values to avoid merging control-flow information
along arms of conditional expressions that can never be taken.
In our experience, the key to reducing the runtime of control-flow analysis while still
maintaining high precision lies in carefully choosing (and empirically tuning)
the tracked abstraction of values.

\subsubsection{Tuning the lattice}
Each time we evaluate an expression whose result is bound to a variable, we need
to update the map with a new abstract value that is the result of merging the
old abstract value and the new value given by the analysis.
In theory, if all that we care about in the analysis is the mapping of call
sites to function identifiers, we could use a simple domain for the
value map ($\mathcal{V}$) based on just the powerset of the function identifiers.
Unfortunately, this domain is insufficiently precise in practice because of the presence of
tuples and datatypes.
Furthermore, SML treats all functions has having a single parameter, which means that
function arguments are packed into tuples at call sites and then extracted in the function
body.
Thus, the domain of abstract values needs to support tracking of information as it moves
into and out of more complicated data structures.

We build a lattice over these abstract values using the $\top$ and $\bot$ 
elements as usual, and treating values of \lstinline{TUPLE} and 
\lstinline{LAMBDAS} type as incomparable.
When two \lstinline{LAMBDAS} values are compared, the subset relationship 
provides an ordering.
It is this ordering that allows us to incrementally merge flow information, up
to a finite limit.
The most interesting portion of our implementation is in the merging of
two \lstinline{TUPLE} values.
In the trivial recursive solution, the analysis may fail to terminate because of
the presence of recursive datatypes (\eg{}, on each iteration over a function
that calls the \lstinline{cons} function, we will wrap another \lstinline{TUPLE}
value around the previous value).
In practice for typical Standard ML programs, we have found that limiting the
tracked depth to 5 and then mapping any further additions to $\top$
results in a good balance of performance and precision.

Note that unlike some other analyses, such as sub-zero CFA~\cite{sub-zero-cfa}, we do not limit the
maximum number of tracked functions per variable.
Avoiding this restriction allows us to use the results of our analysis to support
optimizations that can still be performed when multiple functions flow
to the same call site (unlike inlining).
Furthermore, we have found that reducing the number of tracked function variables has no measurable
impact on the runtime of the analysis, but it removes many optimization 
opportunities (\eg, calling convention optimization across a set of common functions).

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End:

### 2015-02-07 icfp15-servo/intro.tex
Initial commit

--- 2015-02-07 icfp15-servo/cfa.tex
@@
 %!TEX root = paper.tex
-%
-\section{Control-Flow Analysis}
-\label{sec:cfa}
-This section provides a brief background on control-flow analysis (CFA) along with an overview 
-of our specific implementation techniques to achieve both acceptable scalability and precision. 
-A more general introduction to control-flow analysis, in particular the 0CFA
-style that we use, is available in the book by Nielson \etal{}~\cite{principles-prog-analysis}.
-For a detailed comparison of modern approaches to control-flow analysis, see Midtgaard's 
-comprehensive survey~\cite{midtgaard-cfa-survey}.
 
-In brief, while many others have implemented control-flow analysis in their
-compilers~\cite{serrano:cfa-paradigm,mlton-cfa,sub-zero-cfa}, our analysis is
-novel in its tracking of a wider range of values --- including boolean values and 
-tuples --- and its lattice coarsening to balance performance and precision.
-
-\subsection{Overview}
-A control-flow analysis computes a finite map from all of the variables in a program 
-to a conservative abstraction of the values that they can take on during the
-execution of the code.
-That is, it computes a finite map 
-\begin{displaymath}
- \mathcal{V} : \mathtt{VarID} \finmap \mathtt{value}
-\end{displaymath}
-where the $\mathtt{value}$ type is defined as a recursive datatype similar to
-that shown in \figref{fig:values}.
-The special $\top$ (\lstinline{TOP}) and $\bot$ (\lstinline{BOT}) elements 
-indicate either all possible values or no known values, respectively. 
-A \lstinline{TUPLE} value handles both the cases of tuples and ML datatype
-representations, which by this point in the compiler have been desugared into
-either raw values or tagged tuples.
-The \lstinline{LAMBDAS} value is used for a set of variable identifiers, all of
-which are guaranteed to be function identifiers.
-The \lstinline{BOOL} value tracks the flow of literal boolean values through
-the program.
-\begin{figure}[t]
-\begin{lstlisting}[basicstyle=\ttfamily\scriptsize\color{\cdColor}]
-  datatype value
-    = TOP
-    | TUPLE of value list
-    | LAMBDAS of CPS.Var.Set.set
-    | BOOL of bool
-    | BOT
-\end{lstlisting}
-\caption{
-  Abstract values.
-}
-\label{fig:values} 
-\end{figure}%
-
-As an example, consider the following code:
-\begin{lstlisting}[basicstyle=\ttfamily\scriptsize\color{\cdColor}]
-let fun double (x) = x+x
-    and apply (f, n) = f(n)
-in
-    apply (double, 2)
-end
-\end{lstlisting}%
-After running CFA on this example, we have
-\begin{displaymath}
-  \begin{array}{rcl}
-    \mathcal{V}(\texttt{f}) & = & \texttt{LAMBDAS(\SET{double})} \\
-    \mathcal{V}(\texttt{n}) & = & \top \\
-    \mathcal{V}(\texttt{x}) & = & \top
-  \end{array}
-\end{displaymath}%
-These results indicate that the variable \lstinline{f} must be bound to the 
-function \lstinline{double}.
-This example and the \lstinline{value} representation in \figref{fig:values} do not
-track numeric values, which is why \lstinline{n} and \lstinline{x} are mapped to $\top$.
-We are planning to track a richer set of values, including datatype-specific values, in the future
-in order to enable optimizations beyond the ones discussed in this paper.
-
-\subsection{Implementation}
-Our CFA implementation is straightforward and similar in spirit to Serrano's~\cite{serrano:cfa-paradigm}.
-We start with an empty map and walk over the intermediate representation of the 
-program.
-At each expression, we update $\mathcal{V}$ by merging the value-flow information
-until we reach a fixed point where the map no longer changes.
-The most interesting difference from Serrano's implementation is that
-we use our tracked boolean values to avoid merging control-flow information
-along arms of conditional expressions that can never be taken.
-In our experience, the key to reducing the runtime of control-flow analysis while still
-maintaining high precision lies in carefully choosing (and empirically tuning)
-the tracked abstraction of values.
-
-\subsubsection{Tuning the lattice}
-Each time we evaluate an expression whose result is bound to a variable, we need
-to update the map with a new abstract value that is the result of merging the
-old abstract value and the new value given by the analysis.
-In theory, if all that we care about in the analysis is the mapping of call
-sites to function identifiers, we could use a simple domain for the
-value map ($\mathcal{V}$) based on just the powerset of the function identifiers.
-Unfortunately, this domain is insufficiently precise in practice because of the presence of
-tuples and datatypes.
-Furthermore, SML treats all functions has having a single parameter, which means that
-function arguments are packed into tuples at call sites and then extracted in the function
-body.
-Thus, the domain of abstract values needs to support tracking of information as it moves
-into and out of more complicated data structures.
-
-We build a lattice over these abstract values using the $\top$ and $\bot$ 
-elements as usual, and treating values of \lstinline{TUPLE} and 
-\lstinline{LAMBDAS} type as incomparable.
-When two \lstinline{LAMBDAS} values are compared, the subset relationship 
-provides an ordering.
-It is this ordering that allows us to incrementally merge flow information, up
-to a finite limit.
-The most interesting portion of our implementation is in the merging of
-two \lstinline{TUPLE} values.
-In the trivial recursive solution, the analysis may fail to terminate because of
-the presence of recursive datatypes (\eg{}, on each iteration over a function
-that calls the \lstinline{cons} function, we will wrap another \lstinline{TUPLE}
-value around the previous value).
-In practice for typical Standard ML programs, we have found that limiting the
-tracked depth to 5 and then mapping any further additions to $\top$
-results in a good balance of performance and precision.
-
-Note that unlike some other analyses, such as sub-zero CFA~\cite{sub-zero-cfa}, we do not limit the
-maximum number of tracked functions per variable.
-Avoiding this restriction allows us to use the results of our analysis to support
-optimizations that can still be performed when multiple functions flow
-to the same call site (unlike inlining).
-Furthermore, we have found that reducing the number of tracked function variables has no measurable
-impact on the runtime of the analysis, but it removes many optimization 
-opportunities (\eg, calling convention optimization across a set of common functions).
+\section{Introduction}
+\label{sec:intro}
 

### 2015-02-07 icfp15-servo/abstract.tex
Abstract

@@
 \begin{abstract}
-Inlining is an optimization that replaces a call to a function with that function's 
-body.
-This optimization not only reduces the overhead of a function call, but can 
-expose additional optimization opportunities to the compiler, such as removing
-redundant operations or unused conditional branches.
-Another optimization, copy propagation, replaces a redundant copy of a still-live 
-variable with the original.
-Copy propagation can reduce the total number of live variables, reducing register
-pressure and memory usage, and possibly
-eliminating redundant memory-to-memory copies.
-In practice, both of these optimizations are implemented in nearly every
-modern compiler.
+The Servo project is an effort started at Mozilla Research to build a new web browser
+engine that takes advantage of both modern parallel hardware and advances in language
+technology that prevent many of the common security errors that plague all of the other
+modern browsers.
+The new language, Rust, enabled us to take on this project in a type-centric way instead of
+the traditional C++ that relies on coding conventions, imprecise static analyses, and
+fuzzers.
+While many languages before have offered some combination of memory safety, parallelism,
+and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
+with the control over both memory and generated code that is required by systems programmers.
+In this paper, \emph{systems} \emph{software} refers to programs that have some combination
+of soft realtime requirements and memory, CPU, and power usage constraints that require the
+programmer to be able to predict these features from their source code --- without being a
+compiler developer.
 
-These two optimizations are practical to implement and effective in first-order 
-languages, but in languages with lexically-scoped first-class functions (aka, closures), these optimizations 
-are not available to code programmed in a higher-order style.
-With higher-order functions, the analysis challenge has been that the 
-environment at the call site must be the same as at the closure capture location, 
-up to the free variables, or the meaning of the program may change.
-Olin Shivers' 1991 dissertation called this family of optimizations 
-\emph{\superbeta} and he proposed one analysis technique, called \emph{reflow},
-to support these optimizations.
-Unfortunately, reflow has proven too expensive to implement in practice.
-Because these higher-order optimizations are not available in functional-language
-compilers, programmers studiously avoid uses of 
-higher-order values that cannot be optimized (particularly in compiler
-benchmarks).
-
-This paper provides the first practical and effective technique for 
-\superbeta (higher-order) inlining and copy propagation, which
-we call unchanged variable analysis.
-We show that this technique is practical by implementing it in the 
-context of a real compiler for an ML-family language and showing 
-that the required analyses have costs below 3\% of the total 
-compilation time.
-This technique's effectiveness is shown through a set of benchmarks
-and example programs, where this analysis exposes additional potential
- optimization sites.
+Servo is currently over 300k lines of code and implements enough of the web to render and
+process many pages, though it is still a far cry from the over 7 million lines of code in
+the Mozilla Firefox browser and its associated libraries.
+However, we believe that we have implemented enough of the web platform to provide an
+early report on the successes, failures, and open problems remaining in Servo, from the
+point of view of programming languages and runtime research.
+It is our hope that this work will spur additional research in new language features
+suitable for a systems language that also scale up to projects the size of a web browser
+engine.
 \end{abstract}

### 2015-02-16 icfp15-servo/abstract.tex
Flesh out the outline a little.

@@
 \begin{abstract}
-The Servo project is an effort started at Mozilla Research to build a new web browser
+Servo is a project started at Mozilla Research to build a new web browser
 engine that takes advantage of both modern parallel hardware and advances in language
@@
 point of view of programming languages and runtime research.
-It is our hope that this work will spur additional research in new language features
-suitable for a systems language that also scale up to projects the size of a web browser
-engine.
+In this work, we show how a language with an affine type system and many syntactic
+features from the ML family of languages can be successfully used to build state-of-the-art
+systems software.
+We also outline several pitfalls encountered along the way and describe some potential
+areas for future research.
 \end{abstract}

### 2015-02-16 icfp15-servo/intro.tex
Flesh out the outline a little.

@@
 \label{sec:intro}
+Servo\footnote{https://github.com/servo/servo} is a new web browser engine whose development
+was started in February of 2013.
+
+Modern web browser engines have a similar complexity level to that of operating systems.
+These engines are expected to execute arbitrary user programs within a safe sandbox, scale 
+across a wide variety of hardware, manage access to shared system resources, and further 
+do all of that while providing good performance and correct behavior
+against sometimes-incomplete standards.
+With the expectation of executing well on mobile devices and parallel hardware, the 
+fundamental design assumptions have significantly changed since all of the modern
+web engines were originally designed.
+
+Servo is a new web browser engine designed to address the major environment and 
+architectural changes over the last decade.
+To address security, we are using a new systems programming language, Rust,
+which adds static type checking to address by far the top cause of security
+issues in modern web browsers --- failure in handling manual memory allocation
+in \Cplusplus{}.
+We scale across a wide variety of hardware by building either data- or task-parallelism, 
+as appropriate, to each of the stages of the Web layout pipeline.
+Finally, we are improving concurrency by reducing the simultaneous access to data
+structures and using a message-passing architecture between components such as the
+JavaScript engine and the rendering engine that is painting graphics to the screen.
+In this experience report, we discuss the design and architecture of a modern web 
+browser engine, show how modern programming language techniques --- many of which
+originated in the functional programming community --- address these design 
+constraints, and also touch on ongoing challenges and areas of research where we
+would welcome additional community input.
 

### 2015-02-16 icfp15-servo/libraries.tex
Initial commit
Flesh out the outline a little.

--- 2015-02-07 icfp15-servo/intro.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Libraries and abstractions}
+\label{sec:libraries}
+
+Many high-level languages provide abstractions over I/O, threading, parallelism, and concurrency.
+This area is one where we started out with much larger aspirations in Rust but ultimately had to pare back,
+for reasons very similar to the reasons that we have avoided GC by default --- it is difficult to provide a
+predictable, fast implementation that works across all platforms of many systems features.
+For example, thread-local storage (TLS) provides variables that are associated with a given thread.
+In the presence of both lightweight (stackless) and full native threads across a variety of architectures, though,
+it becomes difficult to understand what the cost of initialization and storage are for this feature.
+Even lightweight threads in general, when interacting with a built-in scheduler and generic I/O system that
+has to work across a variety of platforms (including Windows!) were too expensive to support in practice and
+had to be removed.
+
+Even some basic sharing, such as hash tables and vectors has proven challenging.
+For example, most web browsers have implementations of vectors that allow instantiation with a default inline size,
+as they have use cases where they create many thousands of vectors, nearly none of which have more than 4~elements.
+In that case, removing the extra pointer indirection --- particularly if the values are of less than pointer size ---
+can be a significant space savings.
+But, these sorts of libraries are more highly optimized for a web browser than for a general-purpose systems language.
+
+\subsection{Concurrency and parallelism}
+
+Concurrency and data parallelism are two places where Rust's type system is very close to perfectly suited.
+In concurrent and (safe) data parallel libraries, the data operated on by distinct threads is also itself distinct.
+Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
+In the case of concurrency this works quite well --- there are often individual objects that need to be owned by
+a single object at a time.
+
+TODO: code sample
+
+Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
+for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.
+
+TODO: code sample
 

### 2015-02-16 icfp15-servo/rust.tex
Initial commit
Flesh out the outline a little.

--- 2015-02-07 icfp15-servo/intro.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Rust}
+\label{sec:rust}
+
+TODO: basic overview
+
+Bunch of features worked well here.
+
+Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially.
+This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.
+
+Monomorphization, as in MLton, is a great way of providing predictable output code to developers.
+
+Whole-program compilation has some negative impacts in compilation speed, but it also allows for more modular development, since we're not as concerned about whether two things are in the same module for optimization purposes or relying on header-like forced inlining model of C++.
+
+Biggest lesson here is by having a large project like Servo, got very early feedback on how language design decisions affected building a large systems project.
+Contrast with Cyclone's attempt to build something very close to C; don't actually need to be for people to be successful.
+Also note that many language changes had very different metrics in Servo than in the Rust compiler --- compilers do not have the same language feature usage as the likely target programs written in Rust.
+
+\subsection{C++ interop}
+Would like to have full C++ inlining.
+
+\subsection{GC integration}
+Would like to be able to plug in our own GC and stack scanning.
 

### 2015-02-16 icfp15-servo/safety.tex
Initial commit
Flesh out the outline a little.

--- 2015-02-07 icfp15-servo/intro.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Unsafe code}
+\label{sec:safety}
+
+There are several categories of code in Servo that cannot be written in the safe subset of Rust.
+
+\subsection{FFI calls}
+
+Interoperating with native C code often requires raw access to memory allocated in Rust or returns
+pointers to memory that was allocated by the C libraries.
+Translations to an from these raw pointers requires the use of unsafe code.
+
+\subsection{Intrusive data structures}
+
+Rust's ownership model assumes that there is a single owner for each piece of data.
+However, many data structures do not follow that model, in order to provide asymptotic
+improvements in access.
+For example, a doubly-linked list contains a back pointer to the previous element to aid
+in traversals in the opposite direction.
+Many optimized hashtable implementations also have both hash-based access to items
+and a linked list of all of the keys or values.
 

### 2015-02-25 icfp15-servo/browsers.tex
Flesh out the outline a little.
seat-of-pants browser execution description draft

@@
 
+Modern web browsers do not just load static pages, but can also handle pages that have similar
+complexity to native applications.
+From application suites such as the Google Apps\footnote{\url{https://apps.google.com}} to games
+based on the Unreal Engine\footnote{\url{https://blog.mozilla.org/blog/2015/02/24/unreal-engine-4-7-binary-release-includes-html5-export-3/}},
+modern browsers have the ability to
+While Servo is designed to handle applications of that complexity, in this section we will
+focus on the subset of the browser design that supports loading and displaying a single page.
+
+\subsection{Parsing HTML and CSS}
+
+A URL identifies a resource to load.
+This resource usually consists of HTML, which is then parsed and typically turned into a Document Object
+Model (DOM) tree.
+From a programming languages standpoint, there are several interesting aspects of the parser design
+for HTML.
+First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/syntax.html#parsing}},
+in practice browsers follow the recovery algorithms described in that specification and make a best attempt to
+load any page given to them.
+Second, due to the presence of the \lstinline[language=HTML]{<script>} tag, the token stream can be modified
+during operation.
+For example, the below example that injects a close tag for the script block works in all modern browsers.
+\begin{lstlisting}[language=HTML]
+  <html>
+    <script>
+      document.write (``</script>
+    <li>this shows up just fine, thanks</li>
+  </html>
+\end{lstlisting}
+Finally, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
+all modern parsers also perform token stream scanning and prefectch of resources likely to be required. CITE
+
+\subsection{Layout}
+
+After constructing the DOM and determining the set of styles defined in the Cascading Style Sheets (CSS) that has
+been loaded, those styles are applied to the DOM and a new tree that represents the elements for display is
+created.
+This process can create many more elements than previously existed in the DOM --- for example, when a list item is
+styled to have an associated counter glyph.
+
+\subsection{Painting}
+
+Once all of the elements to appear on screen have been computed, these elements are painted into memory buffers or
+directly to graphics surfaces.
+The order of painting these elements is well-defined by the standard\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}.
+
+\subsection{Scripting}
+
+Whether through timers, \lstinline[language=HTML]{<script>} blocks in the HTML, user interactions, or other event handlers,
+JavaScript (JS) code may execute at any point during parsing, layout, and painting or afterwards during display.
+These scripts can modify the DOM tree, which may require rerunning the layout and painting passes in order to update
+the output.
+Most modern browsers use some form of dirty bit marking to attempt to minimize the recalculations during this process.
+
 %%% Local Variables: 

### 2015-02-27 icfp15-servo/abstract.tex
Add a pic. Clean up the abstract a bit

@@
 \begin{abstract}
-Servo is a project started at Mozilla Research to build a new web browser
-engine that takes advantage of both modern parallel hardware and advances in language
-technology that prevent many of the common security errors that plague all of the other
-modern browsers.
-The new language, Rust, enabled us to take on this project in a type-centric way instead of
-the traditional C++ that relies on coding conventions, imprecise static analyses, and
-fuzzers.
-While many languages before have offered some combination of memory safety, parallelism,
-and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
-with the control over both memory and generated code that is required by systems programmers.
-In this paper, \emph{systems} \emph{software} refers to programs that have some combination
-of soft realtime requirements and memory, CPU, and power usage constraints that require the
-programmer to be able to predict these features from their source code --- without being a
-compiler developer.
+All modern web browsers --- Internet Explorer, Firefox, Chrome, Opera, and Webkit ---
+have a core rendering engine written in \Cplusplus.
+This language choice was made because it affords the systems programmer complete control
+of the underlying hardware features and memory in use plus provides a transparent
+compilation model.
 
-Servo is currently over 300k lines of code and implements enough of the web to render and
-process many pages, though it is still a far cry from the over 7 million lines of code in
-the Mozilla Firefox browser and its associated libraries.
-However, we believe that we have implemented enough of the web platform to provide an
-early report on the successes, failures, and open problems remaining in Servo, from the
-point of view of programming languages and runtime research.
-In this work, we show how a language with an affine type system and many syntactic
-features from the ML family of languages can be successfully used to build state-of-the-art
-systems software.
+Servo is a project started at Mozilla Research to build a new web browser engine that
+preserves the capabilities of these other browser engines but also both takes more
+advantage of the recent trends in parallel hardware and is more memory-safe.
+We use a new language, Rust, that provides us a similar level of control of the underlying
+system but which builds on many concepts familiar to the functional programming community
+to form a novelty - a useful, safe systems programming language.
+
+In this paper, we show how a language with an affine type system and many syntactic
+features familiar to functional language programmers can be successfully used to build
+state-of-the-art systems software.
 We also outline several pitfalls encountered along the way and describe some potential

### 2015-02-27 icfp15-servo/browsers.tex
Add a pic. Clean up the abstract a bit

@@
 based on the Unreal Engine\footnote{\url{https://blog.mozilla.org/blog/2015/02/24/unreal-engine-4-7-binary-release-includes-html5-export-3/}},
-modern browsers have the ability to
-While Servo is designed to handle applications of that complexity, in this section we will
-focus on the subset of the browser design that supports loading and displaying a single page.
+modern browsers have the ability to handle much more than simple static pages.
+\figref{fig:browser} shows the steps in processing a site.
+While the naming is specific to the Servo browser, similar steps are used in all modern browsers.\footnote{\url{www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
+\begin{figure*}%[ht]
+  \begin{center}
+    \includegraphics[scale=0.7]{pics/browser}
+  \end{center}%
+  \caption{Processing stages and intermediate representations in a browser engine.}
+  \label{fig:browser}
+\end{figure*}%
 

### 2015-02-27 icfp15-servo/intro.tex
Add a pic. Clean up the abstract a bit

@@
 \label{sec:intro}
+At the heart of a modern web browser is the browser engine, which is the code responsible
+for loading, processing, evaluating, and rendering web content.
+There are three current browser engine families:
+\begin{enumerate}
+\item Trident/Spartan, the engine in Internet Explorer
+\item Webkit/Blink, the engine in Safari, Chrome, and Opera
+\item Gecko, the engine in Firefox
+\end{enumerate}
+All of these engines have at their core many millions of lines of \Cplusplus code.
+While the use of \Cplusplus has enabled all of these browsers to achieve excellent sequential
+performance on a single web page, on mobile devices with lower processor speed but many
+more processors, these browsers do not provide the same level of interactivity that they
+do on desktop processors~\cite{parallelizing-web-pages,ZOOMM}.
+Further, in an informal inspection of the critical security bugs in Gecko, we determined that
+roughly 50\% of the bugs are use after free, out of range access, or related to integer
+overflow.
+The other 50\% is split between errors in tracing values from the JavaScript (JS) heap in the
+\Cplusplus code and errors related to Just-in-time (JIT) code, which we will not discuss
+further in this work.
+
 Servo\footnote{https://github.com/servo/servo} is a new web browser engine whose development
 was started in February of 2013.
-
+The goal of this project is to produce a browser that enables new applications to be authored
+against the web platform that run with more safety, better performance, and better power usage
+than in current browsers.
 Modern web browser engines have a similar complexity level to that of operating systems.
@@
 web engines were originally designed.
+
+In this paper, \emph{systems} \emph{software} refers to programs that have some combination
+of soft realtime requirements and memory, CPU, and power usage constraints that require the
+programmer to be able to predict these features from their source code --- without being a
+compiler developer.
+
+enabled us to take on this project in a type-centric way instead of
+the traditional C++ that relies on coding conventions, imprecise static analyses, and
+fuzzers.
+While many languages before have offered some combination of memory safety, parallelism,
+and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
+with the control over both memory and generated code that is required by systems programmers.
+Servo is currently over 300k lines of code and implements enough of the web to render and
+process many pages, though it is still a far cry from the over 7 million lines of code in
+the Mozilla Firefox browser and its associated libraries.
+However, we believe that we have implemented enough of the web platform to provide an
+early report on the successes, failures, and open problems remaining in Servo, from the
+point of view of programming languages and runtime research.
 

### 2015-02-27 icfp15-servo/rust.tex
Add a pic. Clean up the abstract a bit

--- 2015-02-07 icfp15-servo/intro.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Rust}
+\label{sec:rust}
+
+TODO: basic overview
+
+\subsection{Syntactic features}
+
+Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
+well as pattern matching.
+These types and associated language features provided two large benefits to Servo over traditional browsers
+written in \Cplusplus{}.
+First, providing new abstractions and intermediate representations is syntactically easy, so there is very little
+pressure to tack additional fields into classes simply to avoid creating a large number of new header and implementation
+files.
+More importantly, this style of programming that uses pattern matching over concrete types instead of
+virtual function calls on a class hierarchy provides a non-trivial performance gain.
+Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
+incur indirect function call costs.
+All C++ browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
+specifier wherever possible or specialize the code in some other way to avoid this cost.
+
+Rust also attempted to stay close to familiar syntax, but did not require full fidelity or easy porting of
+programs from languages such as \Cplusplus.
+This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
+with their attempts to build a safe language that preserved most C idioms.
+
+\subsection{Language interoperability}
+Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially.
+This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.
+
+Would like to have full C++ inlining.
+
+\subsection{Compilation strategy}
+Monomorphization, as in MLton, is a great way of providing predictable output code to developers.
+
+Whole-program compilation has some negative impacts in compilation speed, but it also allows for more modular development, since we're not as concerned about whether two things are in the same module for optimization purposes or relying on header-like forced inlining model of C++.
+
+\subsection{Memory management}
+Rust has an ownership model.
+Unlike ML Kit (unbounded regions) and Singularity (ownership bleeding into interfaces), Rust has both statically scoped regions and a notion of borrowing.
+Statically scoped regions
+
+Would like to be able to plug in our own GC and stack scanning.
 

### 2015-02-27 icfp15-servo/browsers.tex
Add related browser research. Cleanups.

@@
 While the naming is specific to the Servo browser, similar steps are used in all modern browsers.\footnote{\url{www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
-\begin{figure*}%[ht]
+\begin{figure*}[ht]
   \begin{center}
@@
 Whether through timers, \lstinline[language=HTML]{<script>} blocks in the HTML, user interactions, or other event handlers,
-JavaScript (JS) code may execute at any point during parsing, layout, and painting or afterwards during display.
+JavaScript code may execute at any point during parsing, layout, and painting or afterwards during display.
 These scripts can modify the DOM tree, which may require rerunning the layout and painting passes in order to update

### 2015-02-27 icfp15-servo/intro.tex
Add related browser research. Cleanups.

@@
 \begin{enumerate}
-\item Trident/Spartan, the engine in Internet Explorer
-\item Webkit/Blink, the engine in Safari, Chrome, and Opera
-\item Gecko, the engine in Firefox
+\item Trident/Spartan, the engine in Internet Explorer~\cite{IE}
+\item Webkit\cite{WEBKIT}/Blink, the engine in Safari~\cite{SAFARI}, Chrome~\cite{CHROME}, and Opera~\cite{OPERA}
+\item Gecko, the engine in Firefox~\cite{FIREFOX}
 \end{enumerate}
@@
 overflow.
-The other 50\% is split between errors in tracing values from the JavaScript (JS) heap in the
-\Cplusplus code and errors related to Just-in-time (JIT) code, which we will not discuss
-further in this work.
+The other 50\% is split between errors in tracing values from the JavaScript heap in the
+\Cplusplus code and errors related to dynamically compiled code.
 

### 2015-02-27 icfp15-servo/open.tex
Initial commit
Add related browser research. Cleanups.

--- 2015-02-07 icfp15-servo/intro.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Open}
+\label{sec:open}
+While this work has discussed many challenges in browser design and our current progress,
+there are many other interesting open problems.
+
+\paragraph{Just-in-time code} is produced from all modern JavaScript engines in order to
+dynamically produce native code that is intended to execute more efficiently than an
+interpreted strategy.
+Unfortunately, this area is a large source of security bugs.
+These bugs come from two sources.
+First, there are potential correctness issues.
+Many of these optimizations are only valid when certain conditions of the calling
+code and environment hold, and ensuring the specialized code is called only when those
+conditions hold is non-trivial to verify.
+Second, dynamically producing and compiling native code and patching it into memory
+while respecting all of the invariants required by the JavaScript runtime (e.g., the
+garbage collector's tracing hooks or free vs. in-use registers) is also a challenge.
+
+\paragraph{JavaScript object rooting} is a problem related to objects from the JavaScript
+heap that are held onto by \Cplusplus object --- and sometimes vice-versa.
+Today, all modern engines manually manage the boundary code to handle tracing
+these objects, and this leads to leaking objects or, worse, freeing objects that
+are still in use.
+The lack of a typical garbage collector for these objects also means that cycles
+between \Cplusplus and JavaScript need to be broken through ad-hoc means.
 

### 2015-02-27 icfp15-servo/rel.tex
Add related browser research. Cleanups.

%!TEX root = paper.tex
%
\section{Related browser research}
\label{sec:rel}
The ZOOMM browser was an effort at Qualcomm Research to build a parallel browser, also
focused on multicore mobile devices~\cite{ZOOMM}.
This browser includes many features that we have not yet implemented in Servo, particularly
around resource fetching.
They also wrote their own JavaScript engine, which we have not done.
Servo and ZOOMM share an extremely concurrent architecture --- both have script, layout,
rendering, and the user interface operating concurrently from one another in order to maximize
interactivity for the user.
Parallel layout in Servo is one major area that was not investigated in the ZOOM browser.
The other major difference is that Servo is implemented in Rust, whereas ZOOM is written
in \Cplusplus, similarly to most modern browser engines.

Ras Bodik's group at the University of California Berkely worked on a parallel browsing
project (funded in part by Mozilla Research) that focused on improving the parallelism
of layout~\cite{parallel-layout}.
Instead of our approach to parallel layout, which focuses on multiple parallel tree
traversals, they modeled a subset of CSS using attribute grammars.
They showed significant speedups with their system over a reimplementation of Safari's
algorithms, but we have not used this approach due to questions of whether it is possible
to use attribute grammars to both accurately model the web as it is implemented today
and to support new features as they are added.
Servo uses a very similar CSS selector matching algorithm to theirs.

### 2015-02-27 icfp15-servo/browsers.tex
More citations instead of footnotes

@@
 From application suites such as the Google Apps\footnote{\url{https://apps.google.com}} to games
-based on the Unreal Engine\footnote{\url{https://blog.mozilla.org/blog/2015/02/24/unreal-engine-4-7-binary-release-includes-html5-export-3/}},
+based on the Unreal Engine\footnote{\url{https://www.unrealengine.com/}},
 modern browsers have the ability to handle much more than simple static pages.

### 2015-02-27 icfp15-servo/intro.tex
More citations instead of footnotes

@@
 \end{enumerate}
-All of these engines have at their core many millions of lines of \Cplusplus code.
-While the use of \Cplusplus has enabled all of these browsers to achieve excellent sequential
+All of these engines have at their core many millions of lines of \Cplusplus{} code.
+While the use of \Cplusplus{} has enabled all of these browsers to achieve excellent sequential
 performance on a single web page, on mobile devices with lower processor speed but many
@@
 The other 50\% is split between errors in tracing values from the JavaScript heap in the
-\Cplusplus code and errors related to dynamically compiled code.
+\Cplusplus{} code and errors related to dynamically compiled code.
 
-Servo\footnote{https://github.com/servo/servo} is a new web browser engine whose development
+Servo~\cite{SERVO} is a new web browser engine whose development
 was started in February of 2013.
@@
 While many languages before have offered some combination of memory safety, parallelism,
-and concurrency, Rust is, to our knowledge, the first to do so that provides the programmer
+and concurrency, Rust~\cite{RUST} is, to our knowledge, the first to do so that provides the programmer
 with the control over both memory and generated code that is required by systems programmers.

### 2015-02-27 icfp15-servo/browsers.tex
More cleanup

@@
 Finally, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
-all modern parsers also perform token stream scanning and prefectch of resources likely to be required. CITE
+all modern parsers also perform token stream scanning and prefectch of resources likely to be required~\cite{browsers-slow-smartphones}.
 
@@
 After constructing the DOM and determining the set of styles defined in the Cascading Style Sheets (CSS) that has
-been loaded, those styles are applied to the DOM and a new tree that represents the elements for display is
-created.
+been loaded, those styles are applied to the DOM and a new tree that represents the elements is created.
 This process can create many more elements than previously existed in the DOM --- for example, when a list item is
@@
 
-\subsection{Painting}
+\subsection{Display}
 
-Once all of the elements to appear on screen have been computed, these elements are painted into memory buffers or
+That tree of elements, called the flow tree in Servo, is then processed to product a set of display list items.
+These list items are the actual graphical elements, text runs, etc. in their final on-screen positions.
+The order in which these elements are displayed is well-defined by the standard\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}.
+
+\subsection{Rendering}
+
+Once all of the elements to appear on screen have been computed, these elements are rendered, or painted, into memory buffers or
 directly to graphics surfaces.
-The order of painting these elements is well-defined by the standard\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}.
 

### 2015-02-27 icfp15-servo/intro.tex
More cleanup

@@
 %!TEX root = paper.tex
-
 \section{Introduction}
 \label{sec:intro}
-At the heart of a modern web browser is the browser engine, which is the code responsible
+The heart of a modern web browser is the browser engine, which is the code responsible
 for loading, processing, evaluating, and rendering web content.
@@
 
-Servo~\cite{SERVO} is a new web browser engine whose development
-was started in February of 2013.
-The goal of this project is to produce a browser that enables new applications to be authored
+Servo~\cite{SERVO} is a new web browser engine designed to address the major environment and 
+architectural changes over the last decade.
+The goal of the Servo project is to produce a browser that enables new applications to be authored
 against the web platform that run with more safety, better performance, and better power usage
 than in current browsers.
-Modern web browser engines have a similar complexity level to that of operating systems.
-These engines are expected to execute arbitrary user programs within a safe sandbox, scale 
-across a wide variety of hardware, manage access to shared system resources, and further 
-do all of that while providing good performance and correct behavior
-against sometimes-incomplete standards.
-With the expectation of executing well on mobile devices and parallel hardware, the 
-fundamental design assumptions have significantly changed since all of the modern
-web engines were originally designed.
-
-In this paper, \emph{systems} \emph{software} refers to programs that have some combination
-of soft realtime requirements and memory, CPU, and power usage constraints that require the
-programmer to be able to predict these features from their source code --- without being a
-compiler developer.
-
-enabled us to take on this project in a type-centric way instead of
-the traditional C++ that relies on coding conventions, imprecise static analyses, and
-fuzzers.
-While many languages before have offered some combination of memory safety, parallelism,
-and concurrency, Rust~\cite{RUST} is, to our knowledge, the first to do so that provides the programmer
-with the control over both memory and generated code that is required by systems programmers.
+To address memory-related safety issues, we are using a new systems programming language,
+Rust~\cite{RUST}.
+For parallelism and power, we scale across a wide variety of hardware by building either data-
+or task-parallelism, as appropriate, to each of the parts of the web platform.
+Additionally, we are improving concurrency by reducing the simultaneous access to data
+structures and using a message-passing architecture between components such as the
+JavaScript engine and the rendering engine that is painting graphics to the screen.
 Servo is currently over 300k lines of code and implements enough of the web to render and
@@
 point of view of programming languages and runtime research.
-
-Servo is a new web browser engine designed to address the major environment and 
-architectural changes over the last decade.
-To address security, we are using a new systems programming language, Rust,
-which adds static type checking to address by far the top cause of security
-issues in modern web browsers --- failure in handling manual memory allocation
-in \Cplusplus{}.
-We scale across a wide variety of hardware by building either data- or task-parallelism, 
-as appropriate, to each of the stages of the Web layout pipeline.
-Finally, we are improving concurrency by reducing the simultaneous access to data
-structures and using a message-passing architecture between components such as the
-JavaScript engine and the rendering engine that is painting graphics to the screen.
 In this experience report, we discuss the design and architecture of a modern web 

### 2015-02-27 icfp15-servo/servo.tex
Complete restructure

\section{Servo}

\subsection{Syntactic features}

Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
well as pattern matching.
These types and associated language features provided two large benefits to Servo over traditional browsers
written in \Cplusplus{}.
First, providing new abstractions and intermediate representations is syntactically easy, so there is very little
pressure to tack additional fields into classes simply to avoid creating a large number of new header and implementation
files.
More importantly, this style of programming that uses pattern matching over concrete types instead of
virtual function calls on a class hierarchy provides a non-trivial performance gain.
Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
incur indirect function call costs.
All C++ browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
specifier wherever possible or specialize the code in some other way to avoid this cost.

Rust also attempted to stay close to familiar syntax, but did not require full fidelity or easy porting of
programs from languages such as \Cplusplus.
This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
with their attempts to build a safe language that preserved most C idioms.

\subsection{Language interoperability}
Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially. This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.

Would like to have full C++ inlining.

\subsection{Compilation strategy}
Monomorphization, as in MLton, is a great way of providing predictable output code to developers.

Whole-program compilation has some negative impacts in compilation speed, but it also allows for more modular development, since we're not as concerned about whether two things are in the same module for optimization purposes or relying on header-like forced inlining model of C++.

\subsection{Memory management}
Rust has an ownership model.
Unlike ML Kit (unbounded regions) and Singularity (ownership bleeding into interfaces), Rust has both statically scoped regions and a notion of borrowing.
Statically scoped regions

Would like to be able to plug in our own GC and stack scanning.

\subsection{Integer overflow/underflow}

\subsection{Affine types}
\label{sec:affine}

Compare with especially Singularity. Key point is having Servo and noticing that most APIs do not transfer ownership. By instead 'borrowing' the value, we do not have region types bleed into the interfaces, and avoid the major failure of Singularity.

\subsection{Unsafe code}
\label{sec:safety}

There are several categories of code in Servo that cannot be written in the safe subset of Rust.

\subsection{FFI calls}

Interoperating with native C code often requires raw access to memory allocated in Rust or returns
pointers to memory that was allocated by the C libraries.
Translations to an from these raw pointers requires the use of unsafe code.

\subsection{Intrusive data structures}

Rust's ownership model assumes that there is a single owner for each piece of data.
However, many data structures do not follow that model, in order to provide asymptotic
improvements in access.
For example, a doubly-linked list contains a back pointer to the previous element to aid
in traversals in the opposite direction.
Many optimized hashtable implementations also have both hash-based access to items
and a linked list of all of the keys or values.
\subsection{Libraries and abstractions}
\label{sec:libraries}

Many high-level languages provide abstractions over I/O, threading, parallelism, and concurrency.
This area is one where we started out with much larger aspirations in Rust but ultimately had to pare back,
for reasons very similar to the reasons that we have avoided GC by default --- it is difficult to provide a
predictable, fast implementation that works across all platforms of many systems features.
For example, thread-local storage (TLS) provides variables that are associated with a given thread.
In the presence of both lightweight (stackless) and full native threads across a variety of architectures, though,
it becomes difficult to understand what the cost of initialization and storage are for this feature.
Even lightweight threads in general, when interacting with a built-in scheduler and generic I/O system that
has to work across a variety of platforms (including Windows!) were too expensive to support in practice and
had to be removed.

Even some basic sharing, such as hash tables and vectors has proven challenging.
For example, most web browsers have implementations of vectors that allow instantiation with a default inline size,
as they have use cases where they create many thousands of vectors, nearly none of which have more than 4~elements.
In that case, removing the extra pointer indirection --- particularly if the values are of less than pointer size ---
can be a significant space savings.
But, these sorts of libraries are more highly optimized for a web browser than for a general-purpose systems language.

\subsection{Concurrency and parallelism}

Concurrency and data parallelism are two places where Rust's type system is very close to perfectly suited.
In concurrent and (safe) data parallel libraries, the data operated on by distinct threads is also itself distinct.
Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
In the case of concurrency this works quite well --- there are often individual objects that need to be owned by
a single object at a time.

%http://is.gd/tkcD9R
\begin{lstlisting}
fn main() {
  let mut data = Box::new(0);

  Thread::spawn(move || {
    *data = *data + 1;
  });
  Thread::spawn(move || {
    *data = *data + 1;
  });
  print!(``{}'', data);
}
\end{lstlisting}

Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.

TODO: code sample

### 2015-02-27 icfp15-servo/open.tex
Some more polish

@@
 
-\paragraph{Just-in-time code} is produced from all modern JavaScript engines in order to
-dynamically produce native code that is intended to execute more efficiently than an
-interpreted strategy.
+\paragraph{Just-in-time code.} JavaScript engines dynamically produce native code that is
+intended to execute more efficiently than an interpreted strategy.
 Unfortunately, this area is a large source of security bugs.
@@
 
-\paragraph{JavaScript object rooting} is a problem related to objects from the JavaScript
-heap that are held onto by \Cplusplus object --- and sometimes vice-versa.
+\paragraph{JavaScript object rooting.} Objects from the JavaScript
+heap  are held onto by \Cplusplus{} object --- and sometimes vice-versa.
 Today, all modern engines manually manage the boundary code to handle tracing
@@
 The lack of a typical garbage collector for these objects also means that cycles
-between \Cplusplus and JavaScript need to be broken through ad-hoc means.
+between \Cplusplus{} and JavaScript need to be broken through ad-hoc means.
+
+\paragraph{Integer overflow/underflow.} It is still an open problem
+to provide optimized code that checks for overflow or underflow without
+incurring significant performance penalties.
+The current plan for Rust is to have debug-only checking of integer ranges
+and for Servo to run debug builds against a test suite, but that may miss
+scenarios that only occur in optimized builds.
 

### 2015-02-27 icfp15-servo/rust.tex
Complete restructure
Some more polish

--- 2015-02-07 icfp15-servo/intro.tex
@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Rust}
+\label{sec:rust}
+
+TODO: basic overview, covering systems language basis; borrowing/ownership \& region/MLKit connection; ADT \& pattern matching; traits \& generics
+
+\subsection{Concurrency and parallelism}
+
+Rust contains library functionality for concurrency and data parallelism.
+Concurrency, in particular, integrates cleanly with Rust's ownership types.
+In safe concurrent programs, the data operated on by distinct threads is also itself distinct.
+Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
+For example, the following code generates a static error from the compiler because after the first
+thread is spawned, the ownership of \lstinline{data} has been transferred into the closure associated
+with that thread and is unavailable for transfer in to the closure for the second thread.
+%http://is.gd/tkcD9R
+\begin{lstlisting}
+fn main() {
+  let mut data = Box::new(0);
+
+  Thread::spawn(move || {
+    *data = *data + 1;
+  });
+  Thread::spawn(move || {
+    *data = *data + 1;
+  });
+  print!(``{}'', data);
+}
+\end{lstlisting}
+
+Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
+for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.
 

### 2015-02-27 icfp15-servo/servo.tex
Some more polish

@@
 \section{Servo}
+\label{sec:servo}
+In this section, we cover specific areas of Servo's design or implementation that make use of Rust
+and the impacts and limitations of these features.
 
-\subsection{Syntactic features}
-
+\subsection{Rust's syntax}
 Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
@@
 Rust also attempted to stay close to familiar syntax, but did not require full fidelity or easy porting of
-programs from languages such as \Cplusplus.
+programs from languages such as \Cplusplus{}.
 This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
 with their attempts to build a safe language that preserved most C idioms.
-
-\subsection{Language interoperability}
-Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially.
-This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.
-
-Would like to have full C++ inlining.
 
@@
 
-\subsection{Integer overflow/underflow}
-
-\subsection{Affine types}
-\label{sec:affine}
-
 Compare with especially Singularity.
@@
 
-\subsection{Unsafe code}
-\label{sec:safety}
+\subsection{Language interoperability}
+Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially.
+This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.
 
-There are several categories of code in Servo that cannot be written in the safe subset of Rust.
+Would like to have full C++ inlining.
 
-\subsection{FFI calls}
-
-Interoperating with native C code often requires raw access to memory allocated in Rust or returns
+That said, interoperating with native C code often requires raw access to memory allocated in Rust or returns
 pointers to memory that was allocated by the C libraries.
-Translations to an from these raw pointers requires the use of unsafe code.
+Translations to and from these raw pointers requires the use of unsafe code.
 
@@
 and a linked list of all of the keys or values.
+
 \subsection{Libraries and abstractions}
-\label{sec:libraries}
 
@@
 But, these sorts of libraries are more highly optimized for a web browser than for a general-purpose systems language.
-
-\subsection{Concurrency and parallelism}
-
-Concurrency and data parallelism are two places where Rust's type system is very close to perfectly suited.
-In concurrent and (safe) data parallel libraries, the data operated on by distinct threads is also itself distinct.
-Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
-In the case of concurrency this works quite well --- there are often individual objects that need to be owned by
-a single object at a time.
-
-%http://is.gd/tkcD9R
-\begin{lstlisting}
-fn main() {
-  let mut data = Box::new(0);
-
-  Thread::spawn(move || {
-    *data = *data + 1;
-  });
-  Thread::spawn(move || {
-    *data = *data + 1;
-  });
-  print!(``{}'', data);
-}
-\end{lstlisting}
-
-Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
-for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.
-
-TODO: code sample

### 2015-02-27 icfp15-servo/browsers.tex
Fix some typos and nits.

@@
 \figref{fig:browser} shows the steps in processing a site.
-While the naming is specific to the Servo browser, similar steps are used in all modern browsers.\footnote{\url{www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
+While the naming is specific to the Servo browser, similar steps are used in all modern browsers.\footnote{\url{http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
 \begin{figure*}[ht]
@@
 for HTML.
-First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/syntax.html#parsing}},
+First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/#parsing}},
 in practice browsers follow the recovery algorithms described in that specification and make a best attempt to
@@
 Finally, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
-all modern parsers also perform token stream scanning and prefectch of resources likely to be required~\cite{browsers-slow-smartphones}.
+all modern parsers also perform token stream scanning and prefetch of resources likely to be required~\cite{browsers-slow-smartphones}.
 
@@
 After constructing the DOM and determining the set of styles defined in the Cascading Style Sheets (CSS) that has
-been loaded, those styles are applied to the DOM and a new tree that represents the elements is created.
-This process can create many more elements than previously existed in the DOM --- for example, when a list item is
+been loaded, those styles are applied to the DOM and a new flow tree that represents the elements is created.
+This process can create many more flows than elements existed in the DOM --- for example, when a list item is
 styled to have an associated counter glyph.
@@
 
-That tree of elements, called the flow tree in Servo, is then processed to product a set of display list items.
+That tree of elements, called the flow tree in Servo, is then processed to produce a set of display list items.
 These list items are the actual graphical elements, text runs, etc. in their final on-screen positions.

### 2015-02-27 icfp15-servo/browsers.tex
Add some XXX comments.

@@
 \end{lstlisting}
+% XXX what is this supposed to show? This is just a script element that contains
+%     the malformed JS |document.write (``|.
 Finally, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),

### 2015-02-27 icfp15-servo/servo.tex
Fix some typos and nits.
Add some XXX comments.

@@
 \subsection{Rust's syntax}
+% XXX Lot's of "provide" here
 Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
 well as pattern matching.
-These types and associated language features provided two large benefits to Servo over traditional browsers
+These types and associated language features provide two large benefits to Servo over traditional browsers
 written in \Cplusplus{}.

### 2015-02-27 icfp15-servo/abstract.tex

@@
 This language choice was made because it affords the systems programmer complete control
-of the underlying hardware features and memory in use plus provides a transparent
+of the underlying hardware features and memory in use and it provides a transparent
 compilation model.

### 2015-02-27 icfp15-servo/intro.tex

@@
 For parallelism and power, we scale across a wide variety of hardware by building either data-
-or task-parallelism, as appropriate, to each of the parts of the web platform.
+or task-parallelism, as appropriate, into each of the parts of the web platform.
 Additionally, we are improving concurrency by reducing the simultaneous access to data

### 2015-02-27 icfp15-servo/rel.tex

@@
 interactivity for the user.
-Parallel layout in Servo is one major area that was not investigated in the ZOOM browser.
-The other major difference is that Servo is implemented in Rust, whereas ZOOM is written
+Parallel layout in Servo is one major area that was not investigated in the ZOOMM browser.
+The other major difference is that Servo is implemented in Rust, whereas ZOOMM is written
 in \Cplusplus, similarly to most modern browser engines.
 
-Ras Bodik's group at the University of California Berkely worked on a parallel browsing
+Ras Bodik's group at the University of California Berkeley worked on a parallel browsing
 project (funded in part by Mozilla Research) that focused on improving the parallelism

### 2015-02-27 icfp15-servo/browsers.tex
Edits from Simon

--- 2015-02-27 icfp15-servo/browsers.tex
@@
 First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/syntax.html#parsing}},
-in practice browsers follow the recovery algorithms described in that specification and make a best attempt to
-load any page given to them.
+in practice browsers follow the recovery algorithms described in that specification precisely so that
+even ill-formed HTML will be handled in an interoperable way across all browsers..
 Second, due to the presence of the \lstinline[language=HTML]{<script>} tag, the token stream can be modified
@@
   <script>
-    document.write (``</script>
-  <li>this shows up just fine, thanks</li>
+  document.write(``<h'');
+  </script>1>
+  This is a h1 title
+
+  <script>
+  document.write(``<!-'');
+  </script>-
+  This is commented
 </html>
 \end{lstlisting}
-Finally, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
-all modern parsers also perform token stream scanning and prefectch of resources likely to be required~\cite{browsers-slow-smartphones}.
+This requirement causes parsing to pause while JavaScript code is running.
+But, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
+all modern parsers also perform token stream scanning and prefetch of resources likely to be required~\cite{browsers-slow-smartphones}.
 

### 2015-02-27 icfp15-servo/rust.tex
Add some XXX comments.
Fix up comment about data parallelism

@@
 
-Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
-for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.
+There is partial support for data parallelism, in the form of partitioning a vector statically in advance
+and transferring ownership of each of those partitions into newly spawned threads.
+But, it is not possible in safe Rust code to write work-stealing algorithms~\cite{blumeofe:multiprogrammed-work-stealing}
+that dynamically parition and adjust work across threads.
 

### 2015-02-27 icfp15-servo/rust.tex
typo

@@
 But, it is not possible in safe Rust code to write work-stealing algorithms~\cite{blumeofe:multiprogrammed-work-stealing}
-that dynamically parition and adjust work across threads.
+that dynamically partition work across threads.
 

### 2015-02-27 icfp15-servo/rust.tex
Rudimentary Rust overview

@@
 
-TODO: basic overview, covering systems language basis; borrowing/ownership \& region/MLKit connection; ADT \& pattern matching; traits \& generics
+Rust is a statically typed systems programming language most heavily inspired by the C
+and ML families of languages~\cite{RUST}.
+Like the C family of languages, it provides the developer control over the use of system
+resources and facilities along with predictable performance.
+In addition, it provides an ownership-based type system inspired by the region systems work in the
+MLKit project~\cite{mlkit} and especially as implemented in the Cyclone project~\cite{cyclone}.
+Unlike the ownership system in the Singularity project~\cite{singularity}, Rust allows APIs to
+not only transfer ownership but also to temporarily \emph{borrow} values, significantly
+reducing the number of region and ownership annotations required in large programs.
+Rust also builds on ideas from typeclasses and type families in Haskell~\cite{haskell}.
+Complete documentation and a language reference for Rust are available at: \url{http://doc.rust-lang.org/}.
 
 \subsection{Concurrency and parallelism}
-
 Rust contains library functionality for concurrency and data parallelism.

### 2015-02-27 icfp15-servo/servo.tex
Clean up the Servo section a bit

--- 2015-02-27 icfp15-servo/servo.tex
@@
 programs from languages such as \Cplusplus{}.
-This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
-with their attempts to build a safe language that preserved most C idioms.
+This approach has worked well for Rust because it has prevented some of the complexity that arose in
+Cyclone~\cite{cyclone} with their attempts to build a safe language that required minimal porting effort
+for even complicated C code.
 
 \subsection{Compilation strategy}
-Monomorphization, as in MLton, is a great way of providing predictable output code to developers.
-
-Whole-program compilation has some negative impacts in compilation speed, but it also allows for more modular development, since we're not as concerned about whether two things are in the same module for optimization purposes or relying on header-like forced inlining model of C++.
+Many statically typed implementations of polymorphic languages such as Standard ML of New Jersey~\cite{SMLNJ} and
+\ocaml{}~\cite{ocaml-manual-3.0} have used a compilation strategy that optimizes representations of data types when
+polymorphic code is monomorphically used, but defaults to a less efficient style otherwise, in order to share
+code~\cite{ocaml-repr}.
+This strategy reduces code size, but leads to unpredictable performance and code, as changes to a codebase that
+either add a new instantiation of a polymorphic function at a given type or, in a modular compilation setting, that
+expose a polymorphic function externally, can change the performance of code that is not local to the change being
+made.
+Monomorphization, as in MLton~\cite{mlton}, instead instantiates each polymorphic code block at each of the types
+it is applied agains, and providing predictable output code to developers, at the cost of some code duplication.
+This approach, combined with whole-program compilation, has some negative impacts in compilation speed but has resulted
+in Rust code that easily matches the sequential speed of its \Cplusplus{} analog without requiring the Servo developers
+to become compiler experts.
 
 \subsection{Memory management}
-Rust has an ownership model.
-Unlike ML Kit (unbounded regions) and Singularity (ownership bleeding into interfaces), Rust has both statically scoped regions and a notion of borrowing.
-Statically scoped regions
+As described in \secref{sec:rust}, Rust has an affine type system that ensures every value is used at
+most once.
+One result of this fact is that in the more than two years since Servo has been under development, we have
+encountered zero use-after-free memory bugs in safe Rust code.
+Given that these bugs make up such a large portion of the security vulnerabilities in modern browsers,
+we believe that even the additional work required to get Rust code to pass the type checker initially is
+justified.
 
-Would like to be able to plug in our own GC and stack scanning.
-
-Compare with especially Singularity.
-Key point is having Servo and noticing that most APIs do not transfer ownership.
-By instead 'borrowing' the value, we do not have region types bleed into the interfaces, and avoid the major failure of Singularity.
+One area for future improvement is related to allocations that are not owned by Rust itself.
+Today, we simple wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
+custom memory alloctor or interoperate with the SpiderMonkey JavaScript engine from Gecko.
+We have implemented wrapper types and compiler plugins that check that these are being used correctly,
+but these are still a source of bugs and one of our largest areas of unsafe code.
 

### 2015-02-27 icfp15-servo/servo.tex
Servo interop section

@@
 made.
-Monomorphization, as in MLton~\cite{mlton}, instead instantiates each polymorphic code block at each of the types
+Monomorphization, as in MLton~\cite{mlton-compiler}, instead instantiates each polymorphic code block at each of the types
 it is applied agains, and providing predictable output code to developers, at the cost of some code duplication.
-This approach, combined with whole-program compilation, has some negative impacts in compilation speed but has resulted
-in Rust code that easily matches the sequential speed of its \Cplusplus{} analog without requiring the Servo developers
-to become compiler experts.
+This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, has a negative impact on
+compilation speed but has resulted in Rust code that easily matches the sequential speed of its \Cplusplus{} analog
+without requiring the Servo developers to become compiler experts.
 
@@
 \subsection{Language interoperability}
-Clean C interop, both for calling and being called were key for integration with various C libraries and bootstrapping larger projects that we intend to eventually be written entirely in Rust, but not initially.
-This support also means being able to define structs with exact layout expected by C for APIs like that, intead of writing shim layers.
+Rust has nearly complete interoperability with C code, both exposing code to and using code from
+C programs.
+This support has allowed us to smoothly integrate with many browser libraries, which has been
+critical for bootstrapping a browser without rewriting all of lower-level libraries immediately, such as
+graphics rendering code, the JavaScript engine, font selection code, etc.
+\tblref{servo-loc} shows the breakdown between current lines of Rust code (including generated code that
+handles interfacing with the JavaScript engine) and C code.
+This table also includes test code, though the majority of that code is in HTML and JavaScript.
+\begin{table}
+  \begin{center}
+    \begin{tabular}{r || r}
+      Language & Lines of Code \\
+      \hline
+      C or \Cplusplus{} & 1,678,427 \\
+      Rust & 410,817 \\
+      HTML or JavaScript & 217,827 \\
+    \end{tabular}%
+  \end{center}%
+  \caption{Lines of code in Servo}
+  \label{servo-loc}
+\end{table}
 
-Would like to have full C++ inlining.
-
-That said, interoperating with native C code often requires raw access to memory allocated in Rust or returns
-pointers to memory that was allocated by the C libraries.
-Translations to and from these raw pointers requires the use of unsafe code.
+There are two limitations in the language interoperability that pose challenges for Servo today.
+First, Rust cannot currently expose varargs-style functions to C code.
+Second, Rust cannot compile against \Cplusplus{} code.
+In both cases, Servo uses C wrapper code to call into the code that Rust cannot directly
+reach.
+While this approach is not a large problem for varargs-style functions, it defeats many of the
+places where the \Cplusplus{} code has been crafted to ensure the code is inlined into the caller,
+which results in degraded performance.
 

### 2015-02-27 icfp15-servo/servo.tex
Move intrusive data structures into memory

@@
 
+Additionally, Rust's ownership model assumes that there is a single owner for each piece of data.
+However, many data structures do not follow that model, in order to provide asymptotic
+improvements in access.
+For example, a doubly-linked list contains a back pointer to the previous element to aid
+in traversals in the opposite direction.
+Many optimized hashtable implementations also have both hash-based access to items
+and a linked list of all of the keys or values.
+In Servo, we have had to use unsafe code to implement data structures with this form,
+though we are typically able to provide a safe interface to users.
+
 \subsection{Language interoperability}
@@
 
-\subsection{Intrusive data structures}
-
-Rust's ownership model assumes that there is a single owner for each piece of data.
-However, many data structures do not follow that model, in order to provide asymptotic
-improvements in access.
-For example, a doubly-linked list contains a back pointer to the previous element to aid
-in traversals in the opposite direction.
-Many optimized hashtable implementations also have both hash-based access to items
-and a linked list of all of the keys or values.
-
 \subsection{Libraries and abstractions}

### 2015-02-27 icfp15-servo/servo.tex
Clean up the libraries subsection

@@
 \subsection{Libraries and abstractions}
-
 Many high-level languages provide abstractions over I/O, threading, parallelism, and concurrency.
@@
 for reasons very similar to the reasons that we have avoided GC by default --- it is difficult to provide a
-predictable, fast implementation that works across all platforms of many systems features.
+predictable, fast implementation that works across all platforms.
 For example, thread-local storage (TLS) provides variables that are associated with a given thread.
@@
 
-Even some basic sharing, such as hash tables and vectors has proven challenging.
-For example, most web browsers have implementations of vectors that allow instantiation with a default inline size,
-as they have use cases where they create many thousands of vectors, nearly none of which have more than 4~elements.
+Therefore, much like other modern browsers, Servo contains many of its own specialized implementations of
+library functions that are tuned for the specific cases of web browsers.
+For example, we have special ``small'' vectors that allow instantiation with a default inline size,
+as there are uses cases where we create many thousands of vectors, nearly none of which have more than 4~elements.
 In that case, removing the extra pointer indirection --- particularly if the values are of less than pointer size ---
 can be a significant space savings.
-But, these sorts of libraries are more highly optimized for a web browser than for a general-purpose systems language.
+We also have our own work-stealing library that has been tuned to work on the flow tree during the process of layout,
+as described in \secref{sec:browsers}.
+It is our hope that this code might be useful to other projects as well, though it is fairly browser-specific today.

### 2015-02-27 icfp15-servo/browsers.tex
More nitpicking

@@
 during operation.
-For example, the below example that injects a close tag for the script block works in all modern browsers.
+For example, the below example that injects an open tag for the header and comment blocks works in all modern browsers.
 \begin{lstlisting}[language=HTML]

### 2015-02-27 icfp15-servo/servo.tex
+me; add a bit about syntax extensions

@@
 It is our hope that this code might be useful to other projects as well, though it is fairly browser-specific today.
+
+\subsection{Compiler plugins}
+
+The Rust compiler has plugin framework that provides a couple of hooks that programmers can use to extend the behavior of the compiler.
+In particular, one can perform arbitrary expansion on the AST and define custom warnings or errors (which also operate on the AST).
+
+We have used this to great effect to interface with the Spidermonkey garbage collector.
+We can enforce, at compile time, that all objects dealing with DOM structs have trace hooks, as well as limiting the amount of ways GC-managed structs can be used to avoid leaking them.
+
+Future plans include hardening our GC safety checks and introducing compile time checks intended to improve performance, for example to reduce redundant cloning of data.

### 2015-02-27 icfp15-servo/servo.tex
More more nitpicking

--- 2015-02-27 icfp15-servo/servo.tex
@@
 Monomorphization, as in MLton~\cite{mlton-compiler}, instead instantiates each polymorphic code block at each of the types
-it is applied agains, and providing predictable output code to developers, at the cost of some code duplication.
+it is applied against, providing predictable output code to developers at the cost of some code duplication.
 This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, has a negative impact on
@@
 Even lightweight threads in general, when interacting with a built-in scheduler and generic I/O system that
-has to work across a variety of platforms (including Windows!) were too expensive to support in practice and
+has to work across a variety of platforms (including Windows!), were too expensive to support in practice and
 had to be removed.
@@
 For example, we have special ``small'' vectors that allow instantiation with a default inline size,
-as there are uses cases where we create many thousands of vectors, nearly none of which have more than 4~elements.
+as there are use cases where we create many thousands of vectors, nearly none of which have more than 4~elements.
 In that case, removing the extra pointer indirection --- particularly if the values are of less than pointer size ---

### 2015-02-27 icfp15-servo/browsers.tex
Clarify parsing description.

--- 2015-02-27 icfp15-servo/browsers.tex
@@
 \end{lstlisting}
-This requirement causes parsing to pause while JavaScript code is running.
+This requires parsing to pause until JavaScript code has run to completion.
 But, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
-all modern parsers also perform token stream scanning and prefetch of resources likely to be required~\cite{browsers-slow-smartphones}.
+all modern parsers also perform speculative token stream scanning and prefetch of resources likely to be required~\cite{browsers-slow-smartphones}.
 

### 2015-02-27 icfp15-servo/rust.tex
Perf numbers

@@
 Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
-For example, the following code generates a static error from the compiler because after the first
-thread is spawned, the ownership of \lstinline{data} has been transferred into the closure associated
-with that thread and is unavailable for transfer in to the closure for the second thread.
+For example, the code in \figref{fig:bad-concurrency} generates a static error from the compiler because
+after the first thread is spawned, the ownership of \lstinline{data} has been transferred into the
+closure associated with that thread and is unavailable for transfer in to the closure for the second thread.
 %http://is.gd/tkcD9R
+\begin{figure}
 \begin{lstlisting}
@@
 \end{lstlisting}
+  \caption{Code that will not compile because it attempts to access mutable state from two threads.}
+  \label{fig:bad-concurrency}
+\end{figure}
 

### 2015-02-27 icfp15-servo/servo.tex
Perf numbers

--- 2015-02-27 icfp15-servo/servo.tex
@@
 and the impacts and limitations of these features.
+The most important test for a browser, though, is performance --- Servo must be at least as fast
+as other browsers at similar tasks to succeed, even if it provides additional memory safety.
+\tblref{servo-perf} shows a preliminary breakdown of the performance in the layout stage (described
+in \secref{sec:browsers}) of rendering several web sites, taken on a modern Macbook Pro.
+\begin{table}
+  \begin{center}
+    \begin{tabular}{r || r | r | r}
+      Site & Gecko & Servo 1 thread & Servo 4 threads \\
+      \hline
+      Reddit & 250 & 100 & 55  \\
+      CNN & 105 & 50 & 35 \\
+    \end{tabular}%
+  \end{center}%
+  \caption{Performance of Servo against Mozilla's Gecko rendering engine on the layout portion of some common sites.
+  Times are in milliseconds.}
+  \label{servo-perf}
+\end{table}
 

### 2015-02-27 icfp15-servo/intro.tex

--- 2015-02-27 icfp15-servo/intro.tex
@@
 For parallelism and power, we scale across a wide variety of hardware by building either data-
-or task-parallelism, as appropriate, to each of the parts of the web platform.
+or task-parallelism, as appropriate, into each part of the web platform.
 Additionally, we are improving concurrency by reducing the simultaneous access to data
 structures and using a message-passing architecture between components such as the
-JavaScript engine and the rendering engine that is painting graphics to the screen.
+JavaScript engine and the rendering engine that paints graphics to the screen.
 Servo is currently over 300k lines of code and implements enough of the web to render and

### 2015-02-27 icfp15-servo/browsers.tex
ms2ger's changes

@@
 for HTML.
-First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/syntax.html#parsing}},
+First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/#parsing}},
 in practice browsers follow the recovery algorithms described in that specification precisely so that
@@
 After constructing the DOM and determining the set of styles defined in the Cascading Style Sheets (CSS) that has
-been loaded, those styles are applied to the DOM and a new tree that represents the elements is created.
-This process can create many more elements than previously existed in the DOM --- for example, when a list item is
+been loaded, those styles are applied to the DOM and a new flow tree that represents the elements is created.
+This process can create many more flows than previously existed in the DOM --- for example, when a list item is
 styled to have an associated counter glyph.

### 2015-02-27 icfp15-servo/servo.tex
ms2ger's changes
ms2ger feedback

@@
 \subsection{Rust's syntax}
-Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
+Rust has struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
 well as pattern matching.
-These types and associated language features provided two large benefits to Servo over traditional browsers
+These types and associated language features provide two large benefits to Servo over traditional browsers
 written in \Cplusplus{}.
-First, providing new abstractions and intermediate representations is syntactically easy, so there is very little
+First, creating new abstractions and intermediate representations is syntactically easy, so there is very little
 pressure to tack additional fields into classes simply to avoid creating a large number of new header and implementation
 files.
-More importantly, this style of programming that uses pattern matching over concrete types instead of
-virtual function calls on a class hierarchy provides a non-trivial performance gain.
+More importantly, pattern matching with static dispatch is typically faster than a virtual function call on a class
+hierarchy.
 Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly

### 2015-02-27 icfp15-servo/servo.tex
Memory management clarifications

--- 2015-02-27 icfp15-servo/servo.tex
@@
 One area for future improvement is related to allocations that are not owned by Rust itself.
-Today, we simple wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
-custom memory alloctor or interoperate with the SpiderMonkey JavaScript engine from Gecko.
-We have implemented wrapper types and compiler plugins that check that these are being used correctly,
+Today, we simply wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
+custom memory allocator or interoperate with the SpiderMonkey JavaScript engine from Gecko.
+We have implemented wrapper types and compiler plugins that restrict incorrect uses of these foreign values,
 but these are still a source of bugs and one of our largest areas of unsafe code.

### 2015-02-27 icfp15-servo/servo.tex
mem mgmt. clarifications from jdm

--- 2015-02-27 icfp15-servo/servo.tex
@@
 One area for future improvement is related to allocations that are not owned by Rust itself.
-Today, we simple wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
+Today, we simply wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
 custom memory alloctor or interoperate with the SpiderMonkey JavaScript engine from Gecko.
-We have implemented wrapper types and compiler plugins that check that these are being used correctly,
+We have implemented wrapper types and compiler plugins that restrict incorrect uses of these foreign values,
 but these are still a source of bugs and one of our largest areas of unsafe code.

### 2015-02-27 icfp15-servo/servo.tex

--- 2015-02-27 icfp15-servo/servo.tex
@@
 
-The Rust compiler has plugin framework that provides a couple of hooks that programmers can use to extend the behavior of the compiler.
-In particular, one can perform arbitrary expansion on the AST and define custom warnings or errors (which also operate on the AST).
+The Rust compiler has a plugin framework that allows programmers to modify the behaviour of the compiler. In particular,
+it is possible to define arbitrary AST transformations, as well as custom warnings or errors (which also operate on the AST).
 
-We have used this to great effect to interface with the Spidermonkey garbage collector.
-We can enforce, at compile time, that all objects dealing with DOM structs have trace hooks, as well as limiting the amount of ways GC-managed structs can be used to avoid leaking them.
+These capabilities enable a safer and more correct interface to the SpiderMonkey garbage collector. For example, we
+can enforce at compile time that, during the tracing phase of garbage collection, all Rust objects visible to the GC
+will report all contained pointers to other GC values, avoiding the threat of incorrectly collecting reachable values.
+Furthermore, we restrict the ways our wrappers around SpiderMonkey pointers can be manipulated, thus turning
+potential runtime memory leaks and ownership semantic API mismatches into static compiler errors instead.
 
-Future plans include hardening our GC safety checks and introducing compile time checks intended to improve performance, for example to reduce redundant cloning of data.
+Future plans include refining our safety checks for garbage collected values, such as flagging invalid ownership
+transference, and introducing compile-time checks for constructs that are non-optimal in terms of performance or
+memory usage.

### 2015-02-27 icfp15-servo/browsers.tex
Various cleanup from first read

@@
 in practice browsers follow the recovery algorithms described in that specification precisely so that
-even ill-formed HTML will be handled in an interoperable way across all browsers..
+even ill-formed HTML will be handled in an interoperable way across all browsers.
 Second, due to the presence of the \lstinline[language=HTML]{<script>} tag, the token stream can be modified
@@
 
-That tree of elements, called the flow tree in Servo, is then processed to produce a set of display list items.
+That tree of elements, called the \emph{flow tree} in Servo, is then processed to produce a set of \emph{display list} items.
 These list items are the actual graphical elements, text runs, etc. in their final on-screen positions.

### 2015-02-27 icfp15-servo/open.tex
Fix some typos and nits.
Various cleanup from first read

@@
 \paragraph{JavaScript object rooting.} Objects from the JavaScript
-heap  are held onto by \Cplusplus{} object --- and sometimes vice-versa.
+heap  are held onto by \Cplusplus{} objects --- and sometimes vice-versa.
 Today, all modern engines manually manage the boundary code to handle tracing
@@
 and for Servo to run debug builds against a test suite, but that may miss
-scenarios that only occur in optimized builds.
+scenarios that only occur in optimized builds or that are not represented
+by the test suite.
 

### 2015-02-27 icfp15-servo/rust.tex
Various cleanup from first read

@@
 In addition, it provides an ownership-based type system inspired by the region systems work in the
-MLKit project~\cite{mlkit} and especially as implemented in the Cyclone project~\cite{cyclone}.
-Unlike the ownership system in the Singularity project~\cite{singularity}, Rust allows APIs to
+MLKit project~\cite{mlkit} and especially as implemented in the Cyclone language~\cite{cyclone}.
+Unlike the related ownership system in Singularity OS~\cite{singularity}, Rust allows APIs to
 not only transfer ownership but also to temporarily \emph{borrow} values, significantly

### 2015-02-27 icfp15-servo/servo.tex

@@
 One area for future improvement is related to allocations that are not owned by Rust itself.
-Today, we simple wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
+Today, we simply wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
 custom memory alloctor or interoperate with the SpiderMonkey JavaScript engine from Gecko.
 We have implemented wrapper types and compiler plugins that check that these are being used correctly,
-but these are still a source of bugs and one of our largest areas of unsafe code.
+but they are still a source of bugs and one of our largest areas of unsafe code.
 

### 2015-02-27 icfp15-servo/open.tex
Clarify open JS problems.

@@
 code and environment hold, and ensuring the specialized code is called only when those
-conditions hold is non-trivial to verify.
+conditions hold is non-trivial.
 Second, dynamically producing and compiling native code and patching it into memory
 while respecting all of the invariants required by the JavaScript runtime (e.g., the
-garbage collector's tracing hooks or free vs. in-use registers) is also a challenge.
+garbage collector's read/write barriers or free vs. in-use registers) is also a challenge.
 
 \paragraph{JavaScript object rooting.} Objects from the JavaScript
-heap  are held onto by \Cplusplus{} objects --- and sometimes vice-versa.
+heap are held onto by \Cplusplus{} objects, and often vice-versa as well.
 Today, all modern engines manually manage the boundary code to handle tracing
-these objects, and this leads to leaking objects or, worse, freeing objects that
-are still in use.
-The lack of a typical garbage collector for these objects also means that cycles
-between \Cplusplus{} and JavaScript need to be broken through ad-hoc means.
+these foreign pointers. Any errors in this code lead to leaked object or, worse,
+freeing objects that are still in use.
+The lack of a holistic garbage collector also means that cycles
+between \Cplusplus{} and JavaScript must be broken on an ad-hoc basis, or
+a specialized cycle collector must be integrated that understands cross-language
+references.
 

### 2015-02-27 icfp15-servo/servo.tex

@@
 custom memory alloctor or interoperate with the SpiderMonkey JavaScript engine from Gecko.
-We have implemented wrapper types and compiler plugins that check that these are being used correctly,
+We have implemented wrapper types and compiler plugins that restrict incorrect uses of these foreign values,
 but they are still a source of bugs and one of our largest areas of unsafe code.

### 2015-02-27 icfp15-servo/servo.tex
Concurrency

@@
 Many high-level languages provide abstractions over I/O, threading, parallelism, and concurrency.
-This area is one where we started out with much larger aspirations in Rust but ultimately had to pare back,
-for reasons very similar to the reasons that we have avoided GC by default --- it is difficult to provide a
-predictable, fast implementation that works across all platforms.
-For example, thread-local storage (TLS) provides variables that are associated with a given thread.
-In the presence of both lightweight (stackless) and full native threads across a variety of architectures, though,
-it becomes difficult to understand what the cost of initialization and storage are for this feature.
-Even lightweight threads in general, when interacting with a built-in scheduler and generic I/O system that
-has to work across a variety of platforms (including Windows!), were too expensive to support in practice and
-had to be removed.
-
+Rust provides functionality that addresses each of these concerns, but they are designed as thin
+wrappers over the underlying services, in order to provide a predictable, fast implementation
+that works across all platforms.
 Therefore, much like other modern browsers, Servo contains many of its own specialized implementations of
@@
 It is our hope that this code might be useful to other projects as well, though it is fairly browser-specific today.
+
+Concurrency is available in Rust in the form of CML-style channels~\cite{reppy:cml}, but with a separation
+between the reader and writer ends of the channel.
+This separation allows Rust to enforce a multiple-writer, single-reader constraint, both simplifying and improving
+the performance of the implementation over one that supports multiple readers.
+We have structured the entire Servo browser engine as a series of threads that communicate over channels,
+avoiding unsafe explicitly shared global memory for all but a single case (reading properties in the flow tree
+from script, an operation whose performance is crucially tested in many browser benchmarks).
+The major challenge we have encountered with this approach is the same one we have heard from other designers
+of large concurrent systems --- reasoning about whether protocols make progress or threads eventually terminate
+is manual and quite challenging, particularly in the presence of arbitrary thread failures.
 

### 2015-02-27 icfp15-servo/open.tex

@@
 
-\paragraph{JavaScript object rooting.} Objects from the JavaScript
-heap are held onto by \Cplusplus{} objects, and often vice-versa as well.
-Today, all modern engines manually manage the boundary code to handle tracing
-these foreign pointers. Any errors in this code lead to leaked object or, worse,
-freeing objects that are still in use.
-The lack of a holistic garbage collector also means that cycles
-between \Cplusplus{} and JavaScript must be broken on an ad-hoc basis, or
-a specialized cycle collector must be integrated that understands cross-language
-references.
-
 \paragraph{Integer overflow/underflow.} It is still an open problem

### 2015-02-27 icfp15-servo/servo.tex

@@
 
-\subsection{Compiler plugins}
+\subsection{Macros}
 
-The Rust compiler has a plugin framework that allows programmers to modify the behaviour of the compiler. In particular,
-it is possible to define arbitrary AST transformations, as well as custom warnings or errors (which also operate on the AST).
+Rust provides a hygienic macro system.
+Macros are defined using a declarative, pattern-based syntax~\cite{Kohlbecker:1987:MDS:41625.41632}.
+The macro system has proven invaluable; we have defined more than one hundred macros throughout Servo and its associated libraries.
 
-These capabilities enable a safer and more correct interface to the SpiderMonkey garbage collector. For example, we
-can enforce at compile time that, during the tracing phase of garbage collection, all Rust objects visible to the GC
-will report all contained pointers to other GC values, avoiding the threat of incorrectly collecting reachable values.
-Furthermore, we restrict the ways our wrappers around SpiderMonkey pointers can be manipulated, thus turning
-potential runtime memory leaks and ownership semantic API mismatches into static compiler errors instead.
+\begin{figure}
+\begin{lstlisting}
+match self.state {
+  states::Data => loop {
+    match get_char!(self) {
+      '&'  => go!(self: consume_char_ref),
+      '<'  => go!(self: to TagOpen),
+      '\0' => go!(self: error; emit '\0'),
+      c    => go!(self: emit c),
+\end{lstlisting}
+  \caption{Incremental HTML tokenizer rules, written in a succinct form using macros. Macro invocations are of the form \lstinline{identifier!(...)}.}
+  \label{fig:tokenizer-macros}
+\end{figure}
+
+For example, our HTML tokenizer rules, such as those shown in \figref{fig:tokenizer-macros}, are written in a macro-based domain specific language that closely matches the format of the HTML specification.\footnote{\url{https://html.spec.whatwg.org/multipage/syntax.html#tokenization}} Another macro handles incremental tokenization, so that the state machine can pause at any time to await further input.
+If no next character is available, the \lstinline{get_char!} macro will cause an early return from the function that contains the macro invocation.
+This careful use of non-local control flow, together with the overall expression-oriented style, makes Servo's HTML tokenizer unusually succinct and comprehensible.
+
+The Rust compiler can also load compiler plugins written in Rust.
+These can perform syntactic transformations beyond the capabilities of the hygienic rule-based macros.
+Compiler plugins use unstable internal APIs, so the maintenance burden is high compared to pattern-based macros.
+Nevertheless, Servo uses procedural macros for a number of purposes, including building perfect hash maps at compile time,\footnote{\url{https://github.com/sfackler/rust-phf}} interning string literals, and auto-generating GC trace hooks.
+Despite the exposure to internal compiler APIs, the deep integration with tooling makes procedural macros an attractive alternative to the traditional systems metaprogramming tools of preprocessors and code generators.
+
+\subsection{Project-specific static analysis}
+
+Compiler plugins can also provide ``lint'' checks\footnote{\url{http://doc.rust-lang.org/book/plugins.html#lint-plugins}} that use the same infrastructure as the compiler's built-in warnings.
+This allows project-specific safety or style checks to integrate deeply with the compiler.
+Lint plugins traverse a typechecked AST, and they can be enabled/disabled within any lexical scope, the same way as built-in warnings.
+
+Lint plugins provide some essential guarantees within Servo.
+Because our DOM objects are managed by the JavaScript garbage collector, we must add GC roots for any DOM object we wish to access from Rust code.
+Interaction with a third-party GC written in \Cplusplus{} is well outside the scope of Rust's built-in guarantees, so we bridge the gap with lint plugins.
+These capabilities enable a safer and more correct interface to the SpiderMonkey garbage collector.
+For example, we can enforce at compile time that, during the tracing phase of garbage collection, all Rust objects visible to the GC will report all contained pointers to other GC values, avoiding the threat of incorrectly collecting reachable values.
+Furthermore, we restrict the ways our wrappers around SpiderMonkey pointers can be manipulated, thus turning potential runtime memory leaks and ownership semantic API mismatches into static compiler errors instead.
+
+As the ``lint'' / ``warning'' terminology suggests, these checks may not catch all possible mistakes.
+Ad-hoc extensions to a type system cannot easily guarantee soundness.
+Rather, lint plugins are a lightweight way to catch mistakes deemed particularly common and/or damaging in practice.
+As plugins are ordinary libraries, members of the Rust community can share lint checks that they have found useful.\footnote{\url{https://github.com/Manishearth/rust-clippy}}
 

### 2015-02-27 icfp15-servo/abstract.tex
Fix nits

--- 2015-02-27 icfp15-servo/abstract.tex
@@
 \begin{abstract}
-All modern web browsers --- Internet Explorer, Firefox, Chrome, Opera, and Webkit ---
-have a core rendering engine written in \Cplusplus.
+All modern web browsers --- Internet Explorer, Firefox, Chrome, Opera, and Safari ---
+have a core rendering engine written in \Cplusplus{}.
 This language choice was made because it affords the systems programmer complete control
-of the underlying hardware features and memory in use plus provides a transparent
+of the underlying hardware features and memory in use, and it provides a transparent
 compilation model.
@@
 Servo is a project started at Mozilla Research to build a new web browser engine that
-preserves the capabilities of these other browser engines but also both takes more
+preserves the capabilities of these other browser engines but also both takes 
 advantage of the recent trends in parallel hardware and is more memory-safe.
 We use a new language, Rust, that provides us a similar level of control of the underlying
-system but which builds on many concepts familiar to the functional programming community
-to form a novelty - a useful, safe systems programming language.
+system to \Cplusplus{} but which builds on many concepts familiar to the functional programming
+community, forming a novelty --- a useful, safe systems programming language.
 

### 2015-02-27 icfp15-servo/intro.tex
Fix nits

@@
 for loading, processing, evaluating, and rendering web content.
-There are three current browser engine families:
+There are three major browser engine families:
 \begin{enumerate}
@@
 overflow.
-The other 50\% is split between errors in tracing values from the JavaScript heap in the
+The other 50\% are split between errors in tracing values from the JavaScript heap in the
 \Cplusplus{} code and errors related to dynamically compiled code.
@@
 JavaScript engine and the rendering engine that paints graphics to the screen.
-Servo is currently over 300k lines of code and implements enough of the web to render and
+Servo is currently over 400k lines of Rust code and implements enough of the web to render and
 process many pages, though it is still a far cry from the over 7 million lines of code in

### 2015-02-27 icfp15-servo/servo.tex
Fix nits

@@
 \label{sec:servo}
-In this section, we cover specific areas of Servo's design or implementation that make use of Rust
-and the impacts and limitations of these features.
-The most important test for a browser, though, is performance --- Servo must be at least as fast
+A crucial test of Servo is performance --- Servo must be at least as fast
 as other browsers at similar tasks to succeed, even if it provides additional memory safety.
-\tblref{servo-perf} shows a preliminary breakdown of the performance in the layout stage (described
-in \secref{sec:browsers}) of rendering several web sites, taken on a modern Macbook Pro.
+\tblref{servo-perf} shows a preliminary comparison of the performance of the layout stage (described
+in \secref{sec:browsers}) of rendering several web sites in Mozilla Firefox's Gecko engine compared
+to Servo, taken on a modern Macbook Pro.
 \begin{table}
@@
   \caption{Performance of Servo against Mozilla's Gecko rendering engine on the layout portion of some common sites.
-  Times are in milliseconds.}
+  Times are in milliseconds, where lower numbers are better.}
   \label{servo-perf}
 \end{table}
+In the remainder of this section, we cover specific areas of Servo's design or implementation that make use of
+Rust and the impacts and limitations of these features.
 
@@
 it is applied against, providing predictable output code to developers at the cost of some code duplication.
-This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, has a negative impact on
-compilation speed but has resulted in Rust code that easily matches the sequential speed of its \Cplusplus{} analog
-without requiring the Servo developers to become compiler experts.
+This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, is used in Rust.
+This strategy has a negative impact on compilation speed but has resulted in Rust code that easily matches the
+sequential speed of its \Cplusplus{} analog without requiring the Servo developers to become compiler experts.
 
@@
 Additionally, Rust's ownership model assumes that there is a single owner for each piece of data.
-However, many data structures do not follow that model, in order to provide asymptotic
-improvements in access.
+However, many data structures do not follow that model, in order to provide multiple traversal
+APIs without favoring the performance of one over the other.
 For example, a doubly-linked list contains a back pointer to the previous element to aid
@@
 places where the \Cplusplus{} code has been crafted to ensure the code is inlined into the caller,
-which results in degraded performance.
+resulting in degraded performance.
 
@@
 This allows project-specific safety or style checks to integrate deeply with the compiler.
-Lint plugins traverse a typechecked AST, and they can be enabled/disabled within any lexical scope, the same way as built-in warnings.
+Lint plugins traverse a typechecked abstract syntax tree (AST), and they can be enabled/disabled within any lexical scope, the same way as built-in warnings.
 

### 2015-02-27 icfp15-servo/browsers.tex
Fix code samples.

@@
   <script>
-  document.write(``<h'');
+  document.write("<h");
   </script>1>
@@
   <script>
-  document.write(``<!-'');
+  document.write("<!-");
   </script>-
   This is commented
+  -->
 </html>

### 2015-02-27 icfp15-servo/servo.tex
Remove 'and/or'; 'or' is already non-exclusive.
Add a note about cross-language inlining

@@
 places where the \Cplusplus{} code has been crafted to ensure the code is inlined into the caller,
-resulting in degraded performance.
+resulting in degraded performance. We intend to fix this through cross-language inlining, taking
+advantage of the fact that both \lstinline{rustc} and \lstinline{clang++} can produce output in the
+LLVM\footnote{\url{http://llvm.org}} intermediate representation, which is subject to link-time
+optimization. We have demonstrated this capability at small scale, but have not yet deployed it
+within Servo.
 
@@
 Ad-hoc extensions to a type system cannot easily guarantee soundness.
-Rather, lint plugins are a lightweight way to catch mistakes deemed particularly common and/or damaging in practice.
+Rather, lint plugins are a lightweight way to catch mistakes deemed particularly common or damaging in practice.
 As plugins are ordinary libraries, members of the Rust community can share lint checks that they have found useful.\footnote{\url{https://github.com/Manishearth/rust-clippy}}

### 2015-02-27 icfp15-servo/servo.tex
Added myself as an author and made some minor edits.

@@
 can be a significant space savings.
-We also have our own work-stealing library that has been tuned to work on the flow tree during the process of layout,
+We also have our own work-stealing library that has been tuned to work on the
+DOM and flow trees during the process of styling and layout,
 as described in \secref{sec:browsers}.
@@
 The Rust compiler can also load compiler plugins written in Rust.
-These can perform syntactic transformations beyond the capabilities of the hygienic rule-based macros.
+These can perform syntactic transformations beyond the capabilities of the hygienic pattern-based macros.
 Compiler plugins use unstable internal APIs, so the maintenance burden is high compared to pattern-based macros.

### 2015-02-28 icfp15-servo/browsers.tex
unsmart the single quotes

@@
   <script>
-  document.write("<h");
+  document.write("<h")\';
   </script>1>
@@
 
-That tree of elements, called the \emph{flow tree} in Servo, is then processed to produce a set of \emph{display list} items.
-These list items are the actual graphical elements, text runs, etc. in their final on-screen positions.
-The order in which these elements are displayed is well-defined by the standard\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}.
+That tree of elements, called the \emph{flow tree} in Servo, is then processed
+to produce a set of \emph{display list} items.  These list items are the
+actual graphical elements, text runs, etc. in their final on-screen positions.
+The order in which these elements are displayed is well-defined by the
+standard\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}.
 
@@
 
-Once all of the elements to appear on screen have been computed, these elements are rendered, or painted, into memory buffers or
-directly to graphics surfaces.
+Once all of the elements to appear on screen have been computed, these
+elements are rendered, or painted, into memory buffers or directly to graphics
+surfaces.
+
+\subsection{Compositing}
+
+The set of memory buffers or graphical surfaces, called layers, are then
+transformed and composited together to form a final image for
+presentation.
 
@@
 
-Whether through timers, \lstinline[language=HTML]{<script>} blocks in the HTML, user interactions, or other event handlers,
-JavaScript code may execute at any point during parsing, layout, and painting or afterwards during display.
-These scripts can modify the DOM tree, which may require rerunning the layout and painting passes in order to update
-the output.
-Most modern browsers use some form of dirty bit marking to attempt to minimize the recalculations during this process.
+Whether through timers, \lstinline[language=HTML]{<script>} blocks in the
+HTML, user interactions, or other event handlers, JavaScript code may execute
+at any point during parsing, layout, and painting or afterwards during
+display.  These scripts can modify the DOM tree, which may require rerunning
+the layout and painting passes in order to update the output.  Most modern
+browsers use some form of dirty bit marking to attempt to minimize the
+recalculations during this process.
 

### 2015-02-28 icfp15-servo/abstract.tex
dherman's suggestions for updating the abstract

@@
 
-In this paper, we show how a language with an affine type system and many syntactic
-features familiar to functional language programmers can be successfully used to build
-state-of-the-art systems software.
+In this paper, we show how a language with an affine type system, ownership types,
+regions, and many syntactic features familiar to functional language programmers can be
+successfully used to build state-of-the-art systems software.
 We also outline several pitfalls encountered along the way and describe some potential

### 2015-02-28 icfp15-servo/browsers.tex
Update figure 1.

@@
   \begin{center}
-    \includegraphics[scale=0.7]{pics/browser}
+    \includegraphics[scale=0.5]{pics/browser}
   \end{center}%
@@
 
+\subsection{Styling}
+
+After constructing the DOM, the browser uses the styling information in linked
+CSS files and in the HTML to compute a styled tree of flows and fragments,
+called the \emph{flow tree} in Servo.
+This process can create many more flows than previously existed in the DOM ---
+for example, when a list item is styled to have an associated counter glyph.
+
 \subsection{Layout}
 
-After constructing the DOM and determining the set of styles defined in the Cascading Style Sheets (CSS) that has
-been loaded, those styles are applied to the DOM and a new flow tree that represents the elements is created.
-This process can create many more flows than previously existed in the DOM --- for example, when a list item is
-styled to have an associated counter glyph.
-
-\subsection{Display}
-
-That tree of elements, called the \emph{flow tree} in Servo, is then processed
-to produce a set of \emph{display list} items.  These list items are the
-actual graphical elements, text runs, etc. in their final on-screen positions.
+The flow tree is then processed to produce a set of \emph{display list} items.
+These list items are the actual graphical elements, text runs, etc. in their
+final on-screen positions.
 The order in which these elements are displayed is well-defined by the
@@
 
-The set of memory buffers or graphical surfaces, called layers, are then
+The set of memory buffers or graphical surfaces, called \emph{layers}, are then
 transformed and composited together to form a final image for
-presentation.
+presentation. Layerization is used to optimize interactive transformations
+like scrolling and certain animations.
 

### 2015-02-28 icfp15-servo/servo.tex
Promote LLVM from a footnote to a reference

@@
 advantage of the fact that both \lstinline{rustc} and \lstinline{clang++} can produce output in the
-LLVM\footnote{\url{http://llvm.org}} intermediate representation, which is subject to link-time
-optimization. We have demonstrated this capability at small scale, but have not yet deployed it
-within Servo.
+LLVM intermediate representation~\cite{LLVM}, which is subject to link-time optimization. We have
+demonstrated this capability at small scale, but have not yet deployed it within Servo.
 

### 2015-02-28 icfp15-servo/servo.tex
Added some stuff to Servo section.

--- 2015-02-27 icfp15-servo/servo.tex
@@
 in \secref{sec:browsers}) of rendering several web sites in Mozilla Firefox's Gecko engine compared
-to Servo, taken on a modern Macbook Pro.
+to Servo, taken on a modern MacBook Pro.
 \begin{table}
@@
 files.
+
 More importantly, pattern matching with static dispatch is typically faster than a virtual function call on a class
@@
 specifier wherever possible or specialize the code in some other way to avoid this cost.
+
+Rust's pattern matching syntax is also worthy of note as it makes implementing
+code that comes from specifications read much closer to the specifications
+than code written with typical imperative style. Match patterns in Servo are
+often annotated with references to the sections of the appropriate
+specification to further aid review and debugging.
 
@@
 justified.
+
+Another example is that Rust's memory safety properties prevent iterator
+invalidation in DOM implementations. Since iterators in the DOM implementation
+may trigger execuation of arbitrary JavaScript code, it is important that that
+code cannot somehow invalidate the iteration, which is one of the common
+patterns leading to the previously mentioned browser bugs.
 
@@
 
+\subsection{Derived Traits}
+
+Rust supports automatic derivation of some built-in traits including the
+Encode and Decode traits which do serialization of data structures. Combined
+with being able to choose serialization implementation, this makes
+interprocess communication code very easy to create.
+
+Currently these derived traits are limited to a fixed set, but it is planned
+for this to be extensible to user defined traits. Servo currently works around
+this limitation by using the macro system. For example, the jstraceable macro
+in Servo automatically implements trace hooks for JavaScript integration
+types.
+
 \subsection{Project-specific static analysis}

### 2015-02-28 icfp15-servo/browsers.tex
the identifier `flow tree` was used before it was declared

--- 2015-02-28 icfp15-servo/browsers.tex
@@
 After constructing the DOM and determining the set of styles defined in the Cascading Style Sheets (CSS) that has
-been loaded, those styles are applied to the DOM and a new flow tree that represents the elements is created.
+been loaded, those styles are applied to the DOM and a new tree that represents the elements is created. In Servo, we call this the \emph{flow tree}.
 This process can create many more flows than previously existed in the DOM --- for example, when a list item is
@@
 
-That tree of elements, called the \emph{flow tree} in Servo, is then processed
+The flow tree is then processed
 to produce a set of \emph{display list} items.  These list items are the

### 2015-02-28 icfp15-servo/rust.tex
Fix code samples.

@@
 and ML families of languages~\cite{RUST}.
-Like the C family of languages, it provides the developer control over the use of system
-resources and facilities along with predictable performance.
-In addition, it provides an ownership-based type system inspired by the region systems work in the
+Like the C family of languages, it provides the developer fine control over memory layout
+and predictable performance.
+Unlike with the C family Rust programs are \emph{memory safe} by default,
+only allowing unsafe operations in specially-delineated blocks.
+
+Rust features an \emph{ownership-based} type system inspired by the region systems work in the
 MLKit project~\cite{mlkit} and especially as implemented in the Cyclone language~\cite{cyclone}.
-Unlike the related ownership system in Singularity OS~\cite{singularity}, Rust allows APIs to
-not only transfer ownership but also to temporarily \emph{borrow} values, significantly
+Unlike the related ownership system in Singularity OS~\cite{singularity}, Rust allows programs to
+not only transfer ownership but also to temporarily \emph{borrow} owned values, significantly
 reducing the number of region and ownership annotations required in large programs.
-Rust also builds on ideas from typeclasses and type families in Haskell~\cite{haskell}.
+The ownership model encourages immutability by default while allowing for controlled
+mutation of owned or uniquely-borrowed values.
+
 Complete documentation and a language reference for Rust are available at: \url{http://doc.rust-lang.org/}.
 
-\subsection{Concurrency and parallelism}
-Rust contains library functionality for concurrency and data parallelism.
-Concurrency, in particular, integrates cleanly with Rust's ownership types.
-In safe concurrent programs, the data operated on by distinct threads is also itself distinct.
-Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
+\subsection{Ownership and concurrency}
+Because the Rust type system provides very strong guarantees about memory aliasing,
+most notably that mutable data is unaliased,
+Rust code is memory safe even in concurrent and multithreaded environments,
+but beyond that Rust also ensures \emph{data-race freedom}.
+
+In concurrent programs, the data operated on by distinct threads is also itself distinct:
+under Rust's ownership model, data cannot be owned by two threads at the same time.
 For example, the code in \figref{fig:bad-concurrency} generates a static error from the compiler because
 after the first thread is spawned, the ownership of \lstinline{data} has been transferred into the
-closure associated with that thread and is unavailable for transfer in to the closure for the second thread.
+closure associated with that thread and is no longer available in the original thread.
 %http://is.gd/tkcD9R
@@
 fn main() {
+  // An owned pointer to a heap-allocated
+  // integer
   let mut data = Box::new(0);
@@
   });
-  Thread::spawn(move || {
-    *data = *data + 1;
-  });
+  // error: accessing moved value
   print!(``{}'', data);
@@
 
-There is partial support for data parallelism, in the form of partitioning a vector statically in advance
-and transferring ownership of each of those partitions into newly spawned threads.
-But, it is not possible in safe Rust code to write work-stealing algorithms~\cite{blumeofe:multiprogrammed-work-stealing}
-that dynamically partition work across threads.
+On the other hand, the immutable value in \figref{fig:shared-concurrency} can be borrowed and shared between multiple threads as long as those threads don't outlive the scope of the data, and even mutable values can be shared as long
+as they are owned by a type that preserves the invariant that mutable memory is unaliased, as with the
+mutex is \figref{fig:shared-mutable-concurrency}.
+
+\begin{figure}
+\begin{lstlisting}
+fn main() {
+  // An immutable borrowed pointer to a
+  // stack-allocated integer
+  let data = &1;
+
+  Thread::scoped(|| {
+    println!(``{}'', data);
+  });
+  print!(``{}'', data);
+}
+\end{lstlisting}
+  \caption{Safely reading immutable state from two threads.}
+  \label{fig:shared-concurrency}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}
+fn main() {
+  // A heap allocated integer protected by an
+  // atomically-reference-counted mutex
+  let data = Arc::new(Mutex::new(0));
+  let data2 = data.clone();
+
+  Thread::scoped(move || {
+    *data2.lock().unwrap() = 1;
+  });
+
+  print!(``{}'', *data.lock().unwrap());
+}
+\end{lstlisting}
+  \caption{Safely mutating state from two threads.}
+  \label{fig:shared-mutable-concurrency}
+\end{figure}
+
+With relatively few simple rules, ownership in Rust enables foolproof task parallism,
+but also data parallism, by partitioning vectors and lending mutable references into properly scoped threads.
+Rust's concurrency abstractions are entirely implemented in libraries, and though
+many advanced concurrent patterns such as work-stealing~~\cite{blumeofe:multiprogrammed-work-stealing}
+cannot be implemented in safe Rust, they can usually be encapsulated in a memory-safe interface.
 

### 2015-02-28 icfp15-servo/rust.tex
Fix bad sentence
This fact is restated later

@@
 Because the Rust type system provides very strong guarantees about memory aliasing,
-most notably that mutable data is unaliased,
 Rust code is memory safe even in concurrent and multithreaded environments,

### 2015-02-28 icfp15-servo/open.tex
nit
More open problems

@@
 
-\section{Open}
+\section{Open problems}
 \label{sec:open}
@@
 
+\paragraph{Unsafe code correctness.} Today, when we write unsafe code in Rust
+there is limited validation of memory lifetimes or type safety within that
+code block.
+However, many of our uses of unsafe code are well-behaved pointer lifetime
+or memory translations that cannot be annotated or inferred in Rust.
+We are very interested in additional annotations that would help us prove
+basic properties about our unsafe code, even if these annotations require a
+theorem prover or ILP solver to check.
+
+\paragraph{Incremental computation.} As mentioned in \secref{sec:browsers},
+all modern browsers use some combination of dirty bit marking and incremental
+recomputation heuristics to avoid reprocessing the full page when
+the state is mutated.
+Unfortunately, these heuristics are not only frequently the source of
+performance differences between browsers, but they are also a source of
+correctness bugs.
+A library that provided a form of self adjusting computation suited to
+incremental recomputation of only the visible part of the page, perhaps
+like the work on Adapton~\cite{adapton}, seems like a promising
+approach to addressing this problem.
+
 %%% Local Variables: 

### 2015-02-28 icfp15-servo/servo.tex
Expand on compilation strategy

@@
 \end{table}
+
 In the remainder of this section, we cover specific areas of Servo's design or implementation that make use of
@@
 made.
+
 Monomorphization, as in MLton~\cite{mlton-compiler}, instead instantiates each polymorphic code block at each of the types
 it is applied against, providing predictable output code to developers at the cost of some code duplication.
-This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, is used in Rust.
-This strategy has a negative impact on compilation speed but has resulted in Rust code that easily matches the
-sequential speed of its \Cplusplus{} analog without requiring the Servo developers to become compiler experts.
+This is the strategy used by virtually all \Cplusplus{} compilers to implement templates, so it is proven and well-known
+within systems programming. Rust follows this approach, although it improves on the ergonomics of \Cplusplus{} templates
+by embedding serialized generic function ASTs within ``compiled'' binaries.
+
+Rust also chooses a fairly large default compilation unit size. A Rust \textit{crate} is subject to whole-program
+compilation~\cite{weeks:whole-program-mlton}, and is optimized as a unit. A crate may comprise hundreds of modules,
+which provide namespacing and abstraction. Module dependencies within a crate are allowed to be cyclic.  The distinction
+between crates and modules reflects the fact that computers have grown by a factor of approximately $10^9$ since C was
+designed for the PDP-7, while the human capacity to understand code as a unit has regretfully not kept pace.
+
+The large compilation unit size slows down compilation, and especially diminishes the ability to build code in parallel.
+It has however enabled us to write Rust code that easily matches the sequential speed of its \Cplusplus{} analog, without requiring the Servo developers to become compiler experts.
+Servo contains about 600 modules within 20 crates.
 

### 2015-02-28 icfp15-servo/abstract.tex
Tighten the abstract a little, per kmc

--- 2015-02-27 icfp15-servo/abstract.tex
@@
 
-In this paper, we show how a language with an affine type system and many syntactic
+In this paper, we show how a language with an affine type system, regions, and many syntactic
 features familiar to functional language programmers can be successfully used to build

### 2015-02-28 icfp15-servo/open.tex
Text cleanups

@@
 recomputation heuristics to avoid reprocessing the full page when
-the state is mutated.
+a mutatation is performed.
 Unfortunately, these heuristics are not only frequently the source of
@@
 incremental recomputation of only the visible part of the page, perhaps
-like the work on Adapton~\cite{adapton}, seems like a promising
-approach to addressing this problem.
+based on the Adapton~\cite{adapton} approach, seems promising.
 

### 2015-02-28 icfp15-servo/rust.tex
Text cleanups
De-LaTeX quotes in code. Soften some of the module-related statements.

@@
 and predictable performance.
-Unlike with the C family Rust programs are \emph{memory safe} by default,
+Unlike C programs, Rust programs are \emph{memory safe} by default,
 only allowing unsafe operations in specially-delineated blocks.
@@
   // error: accessing moved value
-  print!(``{}'', data);
+  print!("{}", data);
 }
@@
   Thread::scoped(|| {
-    println!(``{}'', data);
+    println!("{}", data);
   });
-  print!(``{}'', data);
+  print!("{}", data);
 }
@@
 
-  print!(``{}'', *data.lock().unwrap());
+  print!("{}", *data.lock().unwrap());
 }

### 2015-02-28 icfp15-servo/servo.tex
De-LaTeX quotes in code. Soften some of the module-related statements.

@@
 made.
+
 Monomorphization, as in MLton~\cite{mlton-compiler}, instead instantiates each polymorphic code block at each of the types
 it is applied against, providing predictable output code to developers at the cost of some code duplication.
-This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, is used in Rust.
-This strategy has a negative impact on compilation speed but has resulted in Rust code that easily matches the
-sequential speed of its \Cplusplus{} analog without requiring the Servo developers to become compiler experts.
+This strategy is used by virtually all \Cplusplus{} compilers to implement templates, so it is proven and well-known
+within systems programming.
+Rust also follows this approach, although it improves on the ergonomics of \Cplusplus{} templates
+by embedding serialized generic function ASTs within ``compiled'' binaries.
+
+Rust also chooses a fairly large default compilation unit size. A Rust \textit{crate} is subject to whole-program
+compilation~\cite{weeks:whole-program-mlton}, and is optimized as a unit. A crate may comprise hundreds of modules,
+which provide namespacing and abstraction. Module dependencies within a crate are allowed to be cyclic.
+
+The large compilation unit size slows down compilation and especially diminishes the ability to build code in parallel.
+However, it has enabled us to write Rust code that easily matches the sequential speed of its \Cplusplus{} analog, without requiring the Servo developers to become compiler experts.
+Servo contains about 600 modules within 20 crates.
 

### 2015-02-28 icfp15-servo/open.tex
Clarify unsafe code issues, per jdm

@@
 code block.
-However, many of our uses of unsafe code are well-behaved pointer lifetime
-or memory translations that cannot be annotated or inferred in Rust.
+However, many of our uses of unsafe code are well-behaved translations of
+either pointer lifetimes or data representations that cannot be annotated
+or inferred in Rust.
 We are very interested in additional annotations that would help us prove

### 2015-02-28 icfp15-servo/servo.tex
Cplusplus

@@
 incur indirect function call costs.
-All C++ browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
+All \Cplusplus{} browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
 specifier wherever possible or specialize the code in some other way to avoid this cost.

### 2015-02-28 icfp15-servo/rust.tex
Add section on Rust traits

@@
 fn main() {
-  // An immutable borrowed pointer to a
+  // An immutably borrowed pointer to a
   // stack-allocated integer
@@
 
+\subsection{Traits and polymorphism}
+
+Rust supports both static and dynamic parametric polymorphism via \emph{traits},
+which build on ideas from type classes and type families in Haskell~\cite{haskell}.
+
+As with Haskell type classes, traits define functions and types that are associated
+with all implementing types. Function and types in turn can be parameterized over
+generic types that are bounded by traits, as with the trait that defines overloaded
+addition in \figref{fig:fn-trait}
+
+Such calls to associated functions are statically dispatched, so optimize well,
+but Rust also allows types that implement traits to be packaged with a vtable
+so that functions can be resolved dynamically. Because Rust's generics must
+perform as well as C++'s inlined templates, performance-sensitive Rust
+code tends to prefer the static dispatch of traits-as-typeclasses.
+This is true of much of the standard library.
+
+Rust includes several types of closures that are distinguished by whether
+they own or borrow their environment, and these are defined as traits.
+
+\begin{figure}
+\begin{lstlisting}
+pub trait Add<RHS=Self> {
+  type Output;
+
+  fn add(self, rhs: RHS) -> Self::Output;
+}
+
+fn add_pair<T>(pair: (T, T)) -> T::Output
+                             where T: Add {
+  pair.0 + pair.1
+}
+\end{lstlisting}
+  \caption{Rust trait with associated type.}
+  \label{fig:fn-trait}
+\end{figure}
+
 %%% Local Variables: 

### 2015-03-01 icfp15-servo/rust.tex
Fix a couple of typos.
Spotted a couple of typos whilst reading: parallism -> parallelism.

--- 2015-02-28 icfp15-servo/rust.tex
@@
 
-With relatively few simple rules, ownership in Rust enables foolproof task parallism,
-but also data parallism, by partitioning vectors and lending mutable references into properly scoped threads.
+With relatively few simple rules, ownership in Rust enables foolproof task parallelism,
+but also data parallelism, by partitioning vectors and lending mutable references into properly scoped threads.
 Rust's concurrency abstractions are entirely implemented in libraries, and though

### 2015-03-01 icfp15-servo/rust.tex

--- 2015-02-28 icfp15-servo/rust.tex
@@
 as they are owned by a type that preserves the invariant that mutable memory is unaliased, as with the
-mutex is \figref{fig:shared-mutable-concurrency}.
+mutex in \figref{fig:shared-mutable-concurrency}.
 

### 2015-03-03 icfp15-servo/browsers.tex
Text cleanups
De-LaTeX quotes in code. Soften some of the module-related statements.

--- 2015-02-28 icfp15-servo/browsers.tex
@@
 From application suites such as the Google Apps\footnote{\url{https://apps.google.com}} to games
-based on the Unreal Engine\footnote{\url{https://www.unrealengine.com/}},
+based on the Unreal Engine,\footnote{\url{https://www.unrealengine.com/}}
 modern browsers have the ability to handle much more than simple static pages.
@@
   \begin{center}
-    \includegraphics[scale=0.5]{pics/browser}
+    \includegraphics[scale=0.55]{pics/browser}
   \end{center}%
@@
 for HTML.
-First, though the specification allows the browser to abort on a parse error\footnote{\url{https://html.spec.whatwg.org/multipage/#parsing}},
+First, though the specification allows the browser to abort on a parse error,\footnote{\url{https://html.spec.whatwg.org/multipage/#parsing}}
 in practice browsers follow the recovery algorithms described in that specification precisely so that
@@
   <script>
-  document.write("<h")\';
+  document.write("<h");
   </script>1>
@@
 The order in which these elements are displayed is well-defined by the
-standard\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}.
+standard.\footnote{\url{http://www.w3.org/TR/CSS21/zindex.html#painting-order}}
 

### 2015-03-03 icfp15-servo/servo.tex

@@
 Today, we simply wrap raw C pointers in \lstinline[language=Rust]{unsafe} blocks when we need to use a
-custom memory alloctor or interoperate with the SpiderMonkey JavaScript engine from Gecko.
+custom memory allocator or interoperate with the SpiderMonkey JavaScript engine from Gecko.
 We have implemented wrapper types and compiler plugins that restrict incorrect uses of these foreign values,
@@
 This support has allowed us to smoothly integrate with many browser libraries, which has been
-critical for bootstrapping a browser without rewriting all of lower-level libraries immediately, such as
+critical for bootstrapping a browser without rewriting all of the lower-level libraries immediately, such as
 graphics rendering code, the JavaScript engine, font selection code, etc.

### 2015-03-13 icfp15-servo/servo.tex

@@
 hierarchy.
-Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
+Virtual functions can both have an in-memory storage cost associated with the virtual function tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
 incur indirect function call costs.

### 2015-04-21 icfp15-servo/response.txt
Added our reviews and a draft response

We would like to thank the reviewers for their time spent reviewing the paper and providing their feedback. We agree with the presentation-related comments, in particular that there is writing that could be improved and we could use a better presentation of Rust, and will incorporate this feedback in a final draft.

In the following, text preceded by a '-' is from a reviewer's question or comment.

Review 1

- Section 4.2, what does it means that Rust embeds serialized generic function
ASTs within "compiled" binary? What's the point? When and how these ASTs
are later used?

TODO: ANSWER

We agree that the problems of JIT correctness and incremental computation are not specific to Rust; they are problems faced by Servo (and all web browser engines, generically).

Review 2

In a final draft, we would be happy to provide a more detailed breakdown of Servo's usage of unsafe Rust code, categorized by purpose.

It is more ambiguous to quantify the costs of Rust's guarnatees. It is much harder to get a program to compile, but debugging basic memory safety errors is largely eliminated. We also did not go into great depth on our usage of Rust's memory management abstractions because we largely use our own. Because Rust does not support integer type specialization (https://github.com/rust-lang/rfcs/pull/884), many of the web-specific optimizations made to optimize default sizes and layouts of data structures based on common web usage patterns require explicit encodings.

### 2015-04-21 icfp15-servo/response.txt
Explain serialized ASTs in ICFP response

@@
 
-TODO: ANSWER
+Rust serializes ASTs of items with type parameters into libraries so
+that when those items are later instantiated with specific types, the
+compiler can generate machine code that is optimized for those
+instances. The effect is as with C++ requiring templates to be written
+header files. We will clarify.
+
+- Section 5.1, JS JIT, I'm not sure to understand why you consider than JIT
+correctness is a problem faced by Rust, inside the Servo implementation.
+A little bit of elaboration would have helped me to understand the
+problem you want to address here.
 

### 2015-04-22 icfp15-servo/response.txt
Edit response

--- 2015-04-21 icfp15-servo/response.txt
@@
 
-TODO: ANSWER
+Rust implements generic functions by generating optimized machine code for each
+type instantiation. When a library exports a generic function, the compiler
+will include a serialized AST in the "compiled" output to support instantiation
+by library consumers. We will clarify this in the paper.
 
-We agree that the problems of JIT correctness and incremental computation are not specific to Rust; they are problems faced by Servo (and all web browser engines, generically).
+Similarly, C++ libraries often contain monomorphic code that is compiled once,
+in addition to templates that are exposed in header files. Rust improves
+usability by performing this split automatically.
+
+- Section 5.1, JS JIT, I'm not sure to understand why you consider than JIT
+correctness is a problem faced by Rust, inside the Servo implementation.
+A little bit of elaboration would have helped me to understand the
+problem you want to address here.
+
+We agree that JIT correctness is a problem regardless of implementation language.
+Indeed, we mention it due to the history of security holes resulting from mis-optimizations in SpiderMonkey.
+The open problem for Servo is how to obtain Rust-like memory-safety guarantees on the JIT compiler's output.
+We will clarify this in the final draft.
 
@@
 
-It is more ambiguous to quantify the costs of Rust's guarnatees.
-It is much harder to get a program to compile, but debugging basic memory safety errors is largely eliminated.
+It's harder to quantify the cost/benefit tradeoff of Rust's static checking.
+Gecko is deployed on hundreds of millions of diverse end-user systems, with the result that Mozilla engineers spend substantial time tracking down individually-rare memory-safety bugs.
+For this reason, we believe Rust's up-front checking will be especially valuable for a browser engine.
+
 We also did not go into great depth on our usage of Rust's memory management abstractions because we largely use our own.

### 2015-04-24 icfp15-servo/response.txt
Final response

--- 2015-04-21 icfp15-servo/response.txt
@@
 
-TODO: ANSWER
+Rust serializes ASTs of items with type parameters into libraries so
+that when those items are later instantiated with specific types, the
+compiler can generate machine code that is optimized for those
+instances (i.e., monomorphized). The effect is as with C++ requiring
+templates to be written into header files. We will clarify.
 
-We agree that the problems of JIT correctness and incremental computation are not specific to Rust; they are problems faced by Servo (and all web browser engines, generically).
+- Section 5.1, JS JIT, I'm not sure to understand why you consider than JIT
+correctness is a problem faced by Rust, inside the Servo implementation.
+A little bit of elaboration would have helped me to understand the
+problem you want to address here.
+
+We agree that the problems of JIT correctness and incremental computation are not specific to Rust; they are problems faced by Servo (and all other web browser engines, generically).
 
@@
 
-It is more ambiguous to quantify the costs of Rust's guarnatees.
+It is more ambiguous to quantify the costs of Rust's guarantees.
 It is much harder to get a program to compile, but debugging basic memory safety errors is largely eliminated.

### 2015-04-24 icfp15-servo/response.txt
Keegan's changes

--- 2015-04-22 icfp15-servo/response.txt
@@
 
-Similarly, C++ libraries often contain monomorphic code that is compiled once,
-in addition to templates that are exposed in header files. Rust improves
-usability by performing this split automatically.
-
 - Section 5.1, JS JIT, I'm not sure to understand why you consider than JIT
@@
 
-We agree that JIT correctness is a problem regardless of implementation language.
-Indeed, we mention it due to the history of security holes resulting from mis-optimizations in SpiderMonkey.
-The open problem for Servo is how to obtain Rust-like memory-safety guarantees on the JIT compiler's output.
-We will clarify this in the final draft.
+We agree that the problems of JIT correctness and incremental computation are not specific to Rust; they are problems faced by Servo (and all other web browser engines, generically).
 
@@
 It's harder to quantify the cost/benefit tradeoff of Rust's static checking.
-Gecko is deployed on hundreds of millions of diverse end-user systems, with the result that Mozilla engineers spend substantial time tracking down individually-rare memory-safety bugs.
+Firefox is deployed on hundreds of millions of diverse end-user systems, with the result that Mozilla engineers spend substantial time tracking down individually-rare memory-safety bugs.
 For this reason, we believe Rust's up-front checking will be especially valuable for a browser engine.

### 2015-10-11 icse16-servo/intro.tex
Beginning the paper for ICSE16

--- 2015-02-27 icfp15-servo/intro.tex
@@
 JavaScript engine and the rendering engine that paints graphics to the screen.
-Servo is currently over 400k lines of Rust code and implements enough of the web to render and
+Servo is currently over 400k lines of Rust code and implements enough of the web platform to render and
 process many pages, though it is still a far cry from the over 7 million lines of code in

### 2015-10-11 icse16-servo/paper.tex
Beginning the paper for ICSE16

\documentclass[preprint]{../common/acm_proc_article-sp}

\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{url}
\usepackage{graphicx}

\usepackage{tikz}
\usepackage{tkz-graph}
\usetikzlibrary{arrows}
\usetikzlibrary{fit}					% fitting shapes to coordinates
\usetikzlibrary{backgrounds}	% drawing the background after the foreground
\usetikzlibrary{arrows}

\usepackage{color}
\definecolor{Red}{rgb}{0.9,0.0,0.0}  % fixme
\definecolor{Green}{rgb}{0.0,0.4,0.0}
\definecolor{Blue}{rgb}{0.0,0.0,0.9}
\definecolor{DarkBlue}{rgb}{0.0,0.0,0.75}
\definecolor{Midnight}{rgb}{0.0,0.0,0.5}
\definecolor{Purple}{rgb}{0.5,0.0,0.4}
\definecolor{Black}{rgb}{0.0,0.0,0.0}
\definecolor{Yellow}{rgb}{1.0,1.0, 0.25}
\definecolor{Cyan}{rgb}{0.25,1.0, 1.0}

\newcommand{\cdColor}{Black}
\newcommand{\kwColor}{DarkBlue}
\newcommand{\comColor}{Red}

\usepackage{listings}
\usepackage{textcomp}
\input{../common/lstlang-rust}
\lstset{
  basicstyle=\ttfamily\color{\cdColor},
  keywordstyle=\color{\kwColor}\bfseries,
  commentstyle=\color{\comColor}\itshape,
  language=Rust,
  upquote=true
}

\newcommand{\dcfa}{{$\Delta$}{CFA}}
\newcommand{\gcfa}{{$\Gamma$}{CFA}}

\usepackage{algorithm}
\usepackage{algpseudocode}

\input{../common/common-defs}
\newcommand{\superbeta}{Super-\ensuremath{\beta{}} }

\usepackage{../common/code}

\title{Engineering the Servo Web Browser Engine using Rust}

\numberofauthors{6}

\author{
\alignauthor
Brian Anderson\\
       \affaddr{Mozilla Research}\\
       \email{banderson@mozilla.com}
\alignauthor
Lars Bergstrom\\
       \affaddr{Mozilla Research}\\
       \email{larsberg@mozilla.com}
\alignauthor
Manish Goregaokar\\
       \affaddr{Indian Institute of Technology Bombay}\\
       \email{manishg@iitb.ac.in}
\and
\alignauthor
Josh Matthews\\
       \affaddr{Mozilla}\\
       \email{jdm@mozilla.com}
\alignauthor
Keegan McAllister\titlenote{Work performed while employed at Mozilla}\\
       \email{mcallister.keegan@gmail.com}
\alignauthor
Jack Moffitt\\
       \affaddr{Mozilla}\\
       \email{jack@mozilla.com}
\and
\alignauthor
Simon Sapin\\
       \affaddr{Mozilla}\\
       \email{ssapin@mozilla.com}
}

\date{\today}

\begin{document}

\maketitle

%\thispagestyle{empty}

%\sloppy

\input{abstract}

% A category with the (minimum) three required fields
\category{H.5.4}{Information Interfaces and Presentation}{Hypertext/Hypermedia}
%A category including the fourth, optional field follows...
\category{D.2.11}{Software Engineering}{Software Architectures}[languages, patterns]

\keywords{browser engine, Rust, Servo, concurrency, parallelism} % NOT required for Proceedings

\input{intro}
\input{browsers}
\input{rust}
\input{servo}
\input{open}
\input{rel}

\section{Acknowledgments}
The Servo project was started at Mozilla Research, but has benefitted from
contributions from Samsung, Igalia, and hundreds of volunteer community members.
These contributions amount to more than half of the commits to the project,
and we are grateful for both our partners' and volunteers' efforts to make
this project a success.

We would like to thank Robert Clipsham, David Herman, Tim Kuehn, Lindsey Kuper,
Ms2ger, and Ben Striegel for their comments on drafts of this paper.

\bibliographystyle{../common/alpha}
\bibliography{../common/strings-short,../common/manticore,../common/browser}

\end{document}

### 2015-10-18 icse16-servo/abstract.tex
New code count numbers. Talk more about the soft. eng. bits

--- 2015-02-27 icfp15-servo/abstract.tex
@@
 compilation model.
+Unfortunately, this language is both complex and highly susceptible to memory safety
+issues that potentially lead to security holes.
 
@@
 
-In this paper, we show how a language with an affine type system and many syntactic
-features familiar to functional language programmers can be successfully used to build
-state-of-the-art systems software.
+In this paper, we show how a language with an advanced type system can address many of
+the most common security issues and software engineering challenges in other browser
+engines, while still producing code that has the same performance and memory profile
+required by these applications.
 We also outline several pitfalls encountered along the way and describe some potential

### 2015-10-18 icse16-servo/servo.tex
New code count numbers. Talk more about the soft. eng. bits

--- 2015-03-13 icfp15-servo/servo.tex
@@
 C programs.
-This support has allowed us to smoothly integrate with many browser libraries, which has been
+The ability to use C code has allowed us to smoothly integrate with many browser libraries, which has been
 critical for bootstrapping a browser without rewriting all of the lower-level libraries immediately, such as
 graphics rendering code, the JavaScript engine, font selection code, etc.
+Additionally, interoperation with C is required for components of a browser engine such as media decoders
+for DRM content, which are sometimes delivered by vendors as a binary along with a C API.
 \tblref{servo-loc} shows the breakdown between current lines of Rust code (including generated code that
@@
       \hline
-      C or \Cplusplus{} & 1,678,427 \\
-      Rust & 410,817 \\
-      HTML or JavaScript & 217,827 \\
+      C or \Cplusplus{} & 1,187,939 \\
+      Rust & 816,158 \\
+      HTML or JavaScript & 248,768 \\
     \end{tabular}%
   \end{center}%
-  \caption{Lines of code in Servo}
+  \caption{Lines of code in Servo (October 2015)}
   \label{servo-loc}
 \end{table}
+
+In the future, we hope to expose our Rust libraries written for Servo to C, in order to reuse them
+in the Firefox web brower.
+A first potential libraries is our Rust-based URL parser.\footnote{https://github.com/servo/rust-url}
+While URL parsing seems like a simple and secure task, an audit of the public security bugs\footnote{Firefox public security bugs: http://mzl.la/1GkOiIC} shows that all four of the ones that come from URL parsing are related to either indexing out of
+range of a raw memory buffer or using data that had already been freed --- both of which are
+prevented by Rust code.
 

### 2015-10-22 icse16-servo/rust.tex

--- 2015-03-01 icfp15-servo/rust.tex
@@
 
-  Thread::spawn(move || {
+  thread::spawn(move || {
     *data = *data + 1;
@@
 mutex in \figref{fig:shared-mutable-concurrency}.
+It is also possible to share stack-local data across appropriately scoped threads safely without extra cost.\footnote{\url{https://github.com/Kimundi/scoped-threadpool-rs}}
 
@@
 fn main() {
-  // An immutable borrowed pointer to a
-  // stack-allocated integer
-  let data = &1;
+  // An immutable shared pointer
+  // (atomically reference counted)
+  let data = Arc::new(1);
 
-  Thread::scoped(|| {
-    println!("{}", data);
+  thread::spawn(|| {
+    println!("{}", *data);
   });
-  print!("{}", data);
+  print!("{}", *data);
 }
@@
 
-  Thread::scoped(move || {
+  thread::spawn(move || {
     *data2.lock().unwrap() = 1;
@@
 
-With relatively few simple rules, ownership in Rust enables foolproof task parallism,
-but also data parallism, by partitioning vectors and lending mutable references into properly scoped threads.
+With relatively few simple rules, ownership in Rust enables foolproof task parallelism,
+but also data parallelism, by partitioning vectors and lending mutable references across threads.
 Rust's concurrency abstractions are entirely implemented in libraries, and though

### 2015-10-23 icse16-servo/open.tex
Formatting changes

--- 2015-02-28 icfp15-servo/open.tex
@@
 
-\paragraph{Just-in-time code.} JavaScript engines dynamically produce native code that is
+\subsection{Just-in-time code} JavaScript engines dynamically produce native code that is
 intended to execute more efficiently than an interpreted strategy.
@@
 
-\paragraph{Integer overflow/underflow.} It is still an open problem
+\subsection{Integer overflow/underflow} It is still an open problem
 to provide optimized code that checks for overflow or underflow without
@@
 
-\paragraph{Unsafe code correctness.} Today, when we write unsafe code in Rust
+\subsection{Unsafe code correctness} Today, when we write unsafe code in Rust
 there is limited validation of memory lifetimes or type safety within that
@@
 
-\paragraph{Incremental computation.} As mentioned in \secref{sec:browsers},
+\subsection{Incremental computation} As mentioned in \secref{sec:browsers},
 all modern browsers use some combination of dirty bit marking and incremental

### 2015-10-23 icse16-servo/paper.tex
New code count numbers. Talk more about the soft. eng. bits
Enable sloppy
vgosu's comments

@@
-\documentclass[preprint]{../common/acm_proc_article-sp}
+\documentclass[preprint]{../common/sig-alternate}
 
@@
 Josh Matthews\\
-       \affaddr{Mozilla}\\
+       \affaddr{Mozilla Research}\\
        \email{jdm@mozilla.com}
@@
 Jack Moffitt\\
-       \affaddr{Mozilla}\\
+       \affaddr{Mozilla Research}\\
        \email{jack@mozilla.com}
@@
 Simon Sapin\\
-       \affaddr{Mozilla}\\
+       \affaddr{Mozilla Research}\\
        \email{ssapin@mozilla.com}
@@
 
-%\sloppy
+\sloppy
 
@@
 
-We would like to thank Robert Clipsham, David Herman, Tim Kuehn, Lindsey Kuper,
-Ms2ger, and Ben Striegel for their comments on drafts of this paper.
+We would like to thank Robert Clipsham, Valentin Gosu, David Herman, Tim Kuehn,
+Lindsey Kuper, Ms2ger, and Ben Striegel for their comments on drafts of this paper.
 

### 2015-10-23 icse16-servo/servo.tex
vgosu's comments

@@
 hierarchy.
-Virtual functions can both have an in-memory storage cost associated with the virtual function tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
+Virtual functions can have an in-memory storage cost associated with the virtual function tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
 incur indirect function call costs.
@@
 in the Firefox web brower.
-A first potential libraries is our Rust-based URL parser.\footnote{https://github.com/servo/rust-url}
+A first potential library is our Rust-based URL parser.\footnote{https://github.com/servo/rust-url}
 While URL parsing seems like a simple and secure task, an audit of the public security bugs\footnote{Firefox public security bugs: http://mzl.la/1GkOiIC} shows that all four of the ones that come from URL parsing are related to either indexing out of

### 2015-10-23 icse16-servo/browsers.tex
ICFP reviewer comments

--- 2015-03-03 icfp15-servo/browsers.tex
@@
 transformed and composited together to form a final image for
-presentation. Layerization is used to optimize interactive transformations
-like scrolling and certain animations.
+presentation.
+These layers are then used to optimize interactive transformations, such as scrolling
+and certain animations, by only redrawing the buffers that have changed and
+otherwise simply re-composing the surface that has already been rendered into.
 

### 2015-10-23 icse16-servo/intro.tex
ICFP reviewer comments

@@
 All of these engines have at their core many millions of lines of \Cplusplus{} code.
-While the use of \Cplusplus{} has enabled all of these browsers to achieve excellent sequential
-performance on a single web page, on mobile devices with lower processor speed but many
-more processors, these browsers do not provide the same level of interactivity that they
-do on desktop processors~\cite{parallelizing-web-pages,ZOOMM}.
+The use of \Cplusplus{} has enabled all of these browsers to achieve excellent sequential
+performance on a single web page, particularly on desktop computers.
+But, on mobile devices with lower processor speed but many more processors, these browsers
+do not provide the same level of interactivity~\cite{parallelizing-web-pages,ZOOMM}.
 Further, in an informal inspection of the critical security bugs in Gecko, we determined that

### 2015-10-23 icse16-servo/paper.tex
ICFP reviewer comments

@@
 We would like to thank Robert Clipsham, Valentin Gosu, David Herman, Tim Kuehn,
-Lindsey Kuper, Ms2ger, and Ben Striegel for their comments on drafts of this paper.
+Lindsey Kuper, Ms2ger, Ben Striegel, and previous anonymous reviewers for their
+comments on drafts of this paper.
 

### 2015-10-23 icse16-servo/servo.tex
ICFP reviewer comments

@@
 to Servo, taken on a modern MacBook Pro.
+While these measurements only reflect layout performance, they show that Servo is already faster than Gecko with even a single thread, and achieves speedups with multiple threads.
 \begin{table}
@@
 Rust also follows this approach, although it improves on the ergonomics of \Cplusplus{} templates
-by embedding serialized generic function ASTs within ``compiled'' binaries.
+by embedding serialized generic function ASTs within ``compiled'' binaries so that library consumers can instantiate that AST directly instead of re-parsing the \Cplusplus{} header files for the templates.
 
@@
 We have implemented wrapper types and compiler plugins that restrict incorrect uses of these foreign values,
-but they are still a source of bugs and one of our largest areas of unsafe code.
+but they are still a source of bugs and one of our largest areas of unsafe code.\footnote{https://blog.mozilla.org/research/2014/08/26/javascript-servos-only-garbage-collector/}
 

### 2015-10-23 icse16-servo/abstract.tex
Additional ICFP comment

@@
 system to \Cplusplus{} but which builds on many concepts familiar to the functional programming
-community, forming a novelty --- a useful, safe systems programming language.
+community.
 

### 2015-10-23 icse16-servo/intro.tex
It's implicit that the browser includes libraries

@@
 process many pages, though it is still a far cry from the over 7 million lines of code in
-the Mozilla Firefox browser and its associated libraries.
+the Mozilla Firefox browser.
 However, we believe that we have implemented enough of the web platform to provide an

### 2015-10-23 icse16-servo/browsers.tex
Edit intro to browsers section, move some to paper intro.
The browser section is all about architecture. The first sentence
about how its used is out of place.

@@
 
-Modern web browsers do not just load static pages, but can also handle pages that have similar
-complexity to native applications.
-From application suites such as the Google Apps\footnote{\url{https://apps.google.com}} to games
-based on the Unreal Engine,\footnote{\url{https://www.unrealengine.com/}}
-modern browsers have the ability to handle much more than simple static pages.
-\figref{fig:browser} shows the steps in processing a site.
-While the naming is specific to the Servo browser, similar steps are used in all modern browsers.\footnote{\url{http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
+The architecture of all browsers are broadly the same, as demanded be the specifications of
+the web platform and the shared history and wisdom of browsers and their authors.
+As such, the steps Servo uses in \figref{fig:browser} to process a web page
+are similar to those used in all modern browsers.\footnote{\url{http://www.html5rocks.com/en/tutorials/internals/howbrowserswork/}}
 \begin{figure*}[ht]

### 2015-10-23 icse16-servo/browsers.tex
Small edits to parsing. Explain what a flow tree is better.

@@
 Model (DOM) tree.
-From a programming languages standpoint, there are several interesting aspects of the parser design
+From an implementation standpoint, there are two interesting aspects of the parser design
 for HTML.
@@
 This requires parsing to pause until JavaScript code has run to completion.
-But, since resource loading is such a large factor in the latency of loading many webpages (particularly on mobile),
+But since resource loading is such a large factor in the latency of loading many webpages,
 all modern parsers also perform speculative token stream scanning and prefetch of resources likely to be required~\cite{browsers-slow-smartphones}.
@@
 After constructing the DOM, the browser uses the styling information in linked
-CSS files and in the HTML to compute a styled tree of flows and fragments,
-called the \emph{flow tree} in Servo.
-This process can create many more flows than previously existed in the DOM ---
-for example, when a list item is styled to have an associated counter glyph.
+CSS files and in the HTML to compute a styled tree of flows and fragments.
+This \emph{flow tree}, as it is named in Servo, describes the layout of DOM elements on the page,
+and may contain many more flows than previously existed in the DOM.
+For example, when a list item is styled to have an associated bullet, that bullet
+will itself be represented in the flow tree, though it is not part of the DOM.
 

### 2015-10-23 icse16-servo/intro.tex

@@
 \label{sec:intro}
-The heart of a modern web browser is the browser engine, which is the code responsible
+Modern web browsers do not just display static pages, but run web applications
+with complexity similar to native software.
+From application suites such as Google Apps\footnote{\url{https://apps.google.com}} to games
+based on the Unreal Engine,\footnote{\url{https://www.unrealengine.com/}}
+modern browsers are a delivery platform for the types of rich media experiences historically
+tied to single hardware and operating system platforms.
+
+The architecture called upon to realize these experiences was not exactly designed for the task.
+
+TODO: history of web in one sentence.
+
+The heart of a modern web browser is its engine, the code responsible
 for loading, processing, evaluating, and rendering web content.

### 2015-10-23 icse16-servo/abstract.tex
Fix up abstract and intro

@@
 compilation model.
-Unfortunately, this language is both complex and highly susceptible to memory safety
-issues that potentially lead to security holes.
+Unfortunately, this language is complex (especially to new contributors!), challenging to
+write correct parallel code in, and highly susceptible to memory safety issues that
+potentially lead to security holes.
 
@@
 We use a new language, Rust, that provides us a similar level of control of the underlying
-system to \Cplusplus{} but which builds on many concepts familiar to the functional programming
-community.
+system to \Cplusplus{} but which statically prevents many memory safety issues and
+provides direct support for parallelism and concurrency.
 
@@
 required by these applications.
+This language is also quite accessible to new open source contributors and employees,
+even those without a background in \Cplusplus{} or systems programming.
 We also outline several pitfalls encountered along the way and describe some potential
-areas for future research.
+areas for future improvement.
 \end{abstract}

### 2015-10-23 icse16-servo/intro.tex
Fix up abstract and intro

@@
 \label{sec:intro}
+When most web browsers were originally designed, web pages were mostly static.
 Modern web browsers do not just display static pages, but run web applications
@@
 tied to single hardware and operating system platforms.
-
-The architecture called upon to realize these experiences was not exactly designed for the task.
-
-TODO: history of web in one sentence.
+This shift has greatly increased the amount and complexity of code in a modern web engine
+as well as users' expectations around performance and security.
 
@@
 performance on a single web page, particularly on desktop computers.
-But, on mobile devices with lower processor speed but many more processors, these browsers
+But, they all face several challenges:
+\begin{itemize}
+  \item On mobile devices with lower processor speed but many more processors, these browsers
     do not provide the same level of interactivity~\cite{parallelizing-web-pages,ZOOMM}.
-Further, in an informal inspection of the critical security bugs in Gecko, we determined that
-roughly 50\% of the bugs are use after free, out of range access, or related to integer
-overflow.
-The other 50\% are split between errors in tracing values from the JavaScript heap in the
-\Cplusplus{} code and errors related to dynamically compiled code.
+  \item In Gecko, we determined that roughly 50\% of the security critical bugs are memory use after
+    free, array out of range access, or related to integer overflow.
+  \item As the web has become more interactive, the mostly-sequential architecture of these engines
+    has made it challenging to incorporate new features without sacrificing interactivity.
+  \item With the growth in the popularity of other languages at the expense of \Cplusplus{}, both
+    the number of volunteer contributors to the core \Cplusplus{} parts of these browser engine
+    open source codebases has not grown apace with the increase in the size
+    of the codebase.
+\end{itemize}
 
@@
 than in current browsers.
+
 To address memory-related safety issues, we are using a new systems programming language,
 Rust~\cite{RUST}.
+In Rust, errors such as off-by-one array access or memory buffer use after free are
+prevented statically at compile time.
+
 For parallelism and power, we scale across a wide variety of hardware by building either data-
@@
 JavaScript engine and the rendering engine that paints graphics to the screen.
-Servo is currently over 400k lines of Rust code and implements enough of the web platform to render and
+
+With an average of 5 new contributors per week and several volunteers who have turned
+into key members of the project, we believe that Rust has helped Servo to lower the
+barrier to entry in systems programming.
+
+Servo is currently over 800k lines of Rust code and implements enough of the web platform to render and
 process many pages, though it is still a far cry from the over 7 million lines of code in
@@
 In this experience report, we discuss the design and architecture of a modern web 
-browser engine, show how modern programming language techniques --- many of which
-originated in the functional programming community --- address these design 
-constraints, and also touch on ongoing challenges and areas of research where we
-would welcome additional community input.
+browser engine, show how using the Rust programming language has helped us to  address the
+engineering challenges we have encountered when building the browser engine, and also touch on
+open problems and areas of future investigation.
 

### 2015-10-24 icse16-servo/abstract.tex
Abstract & intro fixes

@@
 the most common security issues and software engineering challenges in other browser
-engines, while still producing code that has the same performance and memory profile
-required by these applications.
+engines, while still producing code that has the same performance and memory profile.
 This language is also quite accessible to new open source contributors and employees,

### 2015-10-24 icse16-servo/intro.tex
Abstract & intro fixes

@@
     do not provide the same level of interactivity~\cite{parallelizing-web-pages,ZOOMM}.
-  \item In Gecko, we determined that roughly 50\% of the security critical bugs are memory use after
-    free, array out of range access, or related to integer overflow.
+  \item In Gecko, roughly 50\% of the security critical bugs are memory use after free, array
+    out of range access, or related to integer overflow, all mistakes commonly made by even
+    experienced C++ programmers with access to the best static analysis tools available.
   \item As the web has become more interactive, the mostly-sequential architecture of these engines
     has made it challenging to incorporate new features without sacrificing interactivity.
-  \item With the growth in the popularity of other languages at the expense of \Cplusplus{}, both
+  \item With the growth in the popularity of other languages at the expense of \Cplusplus{}, 
     the number of volunteer contributors to the core \Cplusplus{} parts of these browser engine
@@
 In Rust, errors such as off-by-one array access or memory buffer use after free are
-prevented statically at compile time.
+prevented by the language and its builtin libraries.
 
@@
 early report on the successes, failures, and open problems remaining in Servo, from the
-point of view of programming languages and runtime research.
+point of view of experimenting with the new programming language, Rust.
 In this experience report, we discuss the design and architecture of a modern web 

### 2015-10-24 icse16-servo/open.tex
Much improved Servo section

@@
 garbage collector's read/write barriers or free vs. in-use registers) is also a challenge.
-
-\subsection{Integer overflow/underflow} It is still an open problem
-to provide optimized code that checks for overflow or underflow without
-incurring significant performance penalties.
-The current plan for Rust is to have debug-only checking of integer ranges
-and for Servo to run debug builds against a test suite, but that may miss
-scenarios that only occur in optimized builds or that are not represented
-by the test suite.
 

### 2015-10-24 icse16-servo/servo.tex
Much improved Servo section

@@
 However, it has enabled us to write Rust code that easily matches the sequential speed of its \Cplusplus{} analog, without requiring the Servo developers to become compiler experts.
-Servo contains about 600 modules within 20 crates.
+Servo contains thousands of modules from over 200 crates.
 
@@
 from script, an operation whose performance is crucially tested in many browser benchmarks).
-The major challenge we have encountered with this approach is the same one we have heard from other designers
-of large concurrent systems --- reasoning about whether protocols make progress or threads eventually terminate
-is manual and quite challenging, particularly in the presence of arbitrary thread failures.
+
+Many other systems programs have struggled with the complexity that arises in systems that have
+many concurrent threads.
+This complexity comes in the form of difficulty reasoning about whether a protocol will terminate, whether
+a message will eventually be handled, etc.
+Fortunately, Rust has a library that implements \emph{session types}, which allow expressing a concurrent
+communication protocol in the type system and having the compiler enforce it~\cite{rust-session-types}.
+We are just starting to use this library in Servo to manage the complexity of concurrency.
 
@@
 memory usage.
+
+\subsection{Integer overflow}
+
+While more rare than memory safety bugs, overflowing an integer is still a
+source of some bugs in modern systems software.
+In Rust, integer overflow is checked by default in debug builds, and we
+have caught several potential bugs simply by compiling the code and running
+our tests with debugging enabled in our automation systems.
+
+\subsection{New contributors}
+
+As mentioned in \secref{sec:intro}, Servo gains roughly 5 new contributors per week.
+While part of that is due to tremendous outreach efforts on the part of the team, we believe that the accessibility
+of Rust to non-systems programmers plays a large role as well.
+Almost none of our new contributors come to Servo with \Cplusplus{} experience, and
+nearly a third of our full-time staff members did not program in it professionally
+before beginning work on Servo.
+
+\subsection{Modular development}
+
+Most large systems projects have a \emph{monorepo} design for their source code control (SCC)
+and build system.
+That is, all of the millions of lines of code that make them up live in a single SCC
+system and use a toplevel build system that drives it all in one step.
+This organization is great for developers on the system, as they can make coordinated changes
+to many parts of the system and complete them in one step.
+However, it has meant that many parts of the browser engine, such as the HTML parser,
+CSS selector matching, or URL parsing, could not easily be separated out and reused by other
+projects, so there are separate projects (which do not typically implement quite the same
+standards as the implementations in browsers!) that attempt to provide these facilities to
+other programs.
+
+In Servo, we instead use a \emph{polyrepo} design, which is inherited from Rust's default
+tooling.
+Our source code is stored across many repositories on GitHub\footnote{\url{http://www.github.com}},
+and we use Rust's Cargo\footnote{\url{http://crates.io}} build system to compile, link, and
+manage the dependencies between all of the parts that make up Servo.
+While this design has the downside that coordinated changes spanning multiple parts of the
+system take many steps, it has the advantage that many of Servo's submodules are used and
+contributed to by developers who do not directly work on Servo.

### 2015-10-24 icse16-servo/concl.tex
Add conclusion. Fix typos.

\section{Conclusion}
\label{sec:concl}
In this experience report, we have described our experiences using the new Rust
programming language to develop Servo.
Based on our experiences to date, we would recommend that either any new systems
project or new submodule within a larger C or \Cplusplus{} systems project strongly
consider the use of Rust as an implementation language.

### 2015-10-24 icse16-servo/open.tex
Add conclusion. Fix typos.

@@
 recomputation heuristics to avoid reprocessing the full page when
-a mutatation is performed.
+a mutation is performed.
 Unfortunately, these heuristics are not only frequently the source of

### 2015-10-24 icse16-servo/rust.tex
Formatting changes

@@
 and ML families of languages~\cite{RUST}.
-Like the C family of languages, it provides the developer fine control over memory layout
+Like the C family of languages, it gives the developer fine control over memory layout
 and predictable performance.
 Unlike C programs, Rust programs are \emph{memory safe} by default,
-only allowing unsafe operations in specially-delineated blocks.
+only allowing unsafe operations in specially-delineated \lstinline[language=Rust]{unsafe} blocks.
 
-Rust features an \emph{ownership-based} type system inspired by the region systems work in the
-MLKit project~\cite{mlkit} and especially as implemented in the Cyclone language~\cite{cyclone}.
-Unlike the related ownership system in Singularity OS~\cite{singularity}, Rust allows programs to
-not only transfer ownership but also to temporarily \emph{borrow} owned values, significantly
-reducing the number of region and ownership annotations required in large programs.
-The ownership model encourages immutability by default while allowing for controlled
-mutation of owned or uniquely-borrowed values.
+Beyond performance and safety Rust features the expressive facilities
+of modern high level languages such as generics, algebraic data types,
+pattern matching, and closures.
+Abstractions in Rust are designed to have predictable and minimal
+performance impact, approaching the ideal of 'zero-cost' abstractions.
 
@@
 
+\subsection{Feature overview}
+
+Rust's syntax draws heavily from C++, and many of its features will be
+familiar to modern programmers, though the details of both
+often differ in significant and interesting ways.
+What follows is a brief primer of the basics needed to understand the examples in this paper.
+
+Local variables, declared with \lstinline[language=Rust]{let}, are immutable by default.
+To mutate a variable in Rust it must be declared \lstinline[language=Rust]{mut}.
+Both scalars and aggregates like structs are value types, allocated on the stack.
+The heap is accessed through library container types, the simplest of which are \lstinline[language=Rust]{Box},
+a pointer to a value on the heap, and \lstinline[language=Rust]{Vec}, a dynamically-sized
+array of values. Both mutability and simple heap types are
+illustrated in \figref{fig:mut}.
+
+\begin{figure}
+\begin{lstlisting}
+let max = 100;
+let mut counter = 0;
+let mut boxes: Vec<Box<i32>> = Vec::new();
+while counter < max {
+  // Append a boxed integer to a vector
+  boxes.push(Box::new(counter));
+  counter += 1;
+}
+\end{lstlisting}
+  \caption{Mutability and heap allocation.}
+  \label{fig:mut}
+\end{figure}
+
+Rust's most prominent influence from the ML languages are algebraic data types,
+along with \emph{pattern matching} to destructure
+them into their components. Rust calls algebraic data types \emph{enums},
+and they are the union of multiple types,
+each of which contains their own fields. Although superficially
+similar to C unions, they are significantly different in that
+instances of enums cannot be indiscriminately cast between
+variants; to access the values in an enum one must employ the \lstinline[language=Rust]{match}
+expression to first check the variant, then bind references
+to its interior fields. In \figref{fig:enums} the values of
+a drawing message are extracted from an enum for further processing,
+with the bindings to the enum's fields established to the left of the
+fat arrow (\lstinline[language=Rust]{=>}) and the block of code operating
+on those bindings to the right.
+
+\begin{figure}
+\begin{lstlisting}
+enum Canvas2dMsg {
+  BeginPath,
+  ArcTo(Point2D<f32>, Point2D<f32>, f32),
+  EndPath
+}
+
+let msg = Canvas2dMsg::BeginPath;
+
+match msg {
+  BeginPath => { }
+  EndPath => { }
+  ArcTo(point_a, point_b, radius) => {
+    draw_arc_to(point_a, point_b, radius);
+  }
+}
+\end{lstlisting}
+  \caption{Pattern matching to access the fields of an enum variant.}
+  \label{fig:enums}
+\end{figure}
+
+Both enums and structs (which are substantially similar to C structs)
+may have associated instance methods and static methods, called with the dot
+(\lstinline[language=Rust]{.}) operator and the double-colon
+(\lstinline[language=Rust]{::}) operator, respectively.
+Static methods are common in Rust as they are used to construct
+values, conventionally with a method called
+\lstinline[language=Rust]{new}, as in
+\lstinline[language=Rust]{Ponit2D::new(0.0, 0.0)}.
+Notably, methods in Rust are dispatched statically, making them
+eligable for inlining and aggressive optimization; methods are never
+dispatched dynamically through a function pointer in Rust unless
+explicitly requested.
+
+Like many modern languages, Rust has first-class closures,
+anonymous functions that capture their environment. These
+use a compact notation with the arguments between pipes,
+and the body between brackets: \lstinline[language=Rust]~|a, b| { a + b }~.
+The closure that takes no arguments and performs no computation is
+thus signified by the adorable series of characters, \lstinline[language=Rust]~||{}~.
+Rust closures are translated as efficiently as other types
+in Rust and can be inlined and optimized as well as any function.
+
+Finally, Rust has a hygeinic macro system, much more powerful than
+the C preprocessor. Macro invocations look like function invocations
+where the function name is appended with a bang (\lstinline[language=Rust]{!}).
+They are not functions though, but are expanded in place at compile time,
+and perform syntactic transformations unavailable to functions.
+The common method of performing console output, \lstinline[language=Rust]~println!(``{}'', foo)~,
+is a macro.
+See \figref{fig:bad-concurrency} for an example of closures and macros.
+
+\subsection{Ownership}
+The features of Rust described so far are common to many languages.
+Rust's novelty though is in \emph{ownership and borrowing}.
+In Rust, each value is uniquly owned by a single variable, and
+assigning that value to another transfers ownership of, and access to,
+that value to the receiver. Rust calls this transfer of ownership a \emph{move}.
+When a value is allowed to go out of scope
+it is destroyed. In this way ownership-based resource management is
+like RAII\footnote{``resource acquisition is initialization''} in C++.
+
+To make using variables usably ergonomic Rust can create temporary
+references to values with the ampersand (\lstinline[language=Rust]{&})
+operator. References are simply pointers with limited scope, and may be either
+immutable or mutable. They are the only type of pointer in safe Rust.
+\emph{The rules for borrowing ensure that no two pointers are ever mutably
+aliased, and that invalid memory is never accessible.
+This is what makes Rust memory safe.} See the example in \figref{fig:own}.
+
+\begin{figure}
+\begin{lstlisting}
+
+// The single value we'll work with.
+let val = Box::foo(0);
+
+// Move the value in a to a new location
+let moved_val = val;
+// `val` is no longer accessible.
+// The following will not compile.
+// println(``{}'', val);
+
+// Borrow and share the value immutably
+// several times. This is done in a new
+// block to limit the scope of the borrows.
+{
+  let shared1 = &moved_val;
+  let shared2 = &moved_val;
+  println!(``{}'', shared1);
+  println!(``{}'', shared2);
+  // Borrows are released at the end
+  // of the block.
+}
+\end{lstlisting}
+  \caption{Ownership and borrowing.}
+  \label{fig:own}
+\end{figure}
+
+Ownership-based type systems are a kind of \emph{affine} type system. Rust's
+model is influenced by the ownership model of Singularity OS~\cite{singularity},
+as well as the region systems in the Cyclone language~\cite{cyclone} and MLKit~\cite{mlkit}.
+
 \subsection{Ownership and concurrency}
+
 Because the Rust type system provides very strong guarantees about memory aliasing,
 Rust code is memory safe even in concurrent and multithreaded environments,
-but beyond that Rust also ensures \emph{data-race freedom}.
+and, perhaps surprisingly, is even guaranteed to be \emph{data-race free}.
 
@@
 
+  // The `move` keyword here moves ownership of
+  // the environment into the closure.
   thread::spawn(move || {
@@
 mutex in \figref{fig:shared-mutable-concurrency}.
-It is also possible to share stack-local data across appropriately scoped threads safely without extra cost.\footnote{\url{https://github.com/Kimundi/scoped-threadpool-rs}}
 
@@
   let data = Arc::new(1);
+  let shared_data = data.clone();
 
-  thread::spawn(|| {
-    println!("{}", *data);
+  thread::spawn(move || {
+    println!("{}", *shared_data);
   });
@@
   let data = Arc::new(Mutex::new(0));
-  let data2 = data.clone();
+  let shared_data = data.clone();
 
   thread::spawn(move || {
-    *data2.lock().unwrap() = 1;
+    *shared_data.lock().unwrap() = 1;
   });

### 2015-10-24 icse16-servo/rust.tex
Typos

@@
 \lstinline[language=Rust]{new}, as in
-\lstinline[language=Rust]{Ponit2D::new(0.0, 0.0)}.
+\lstinline[language=Rust]{Point2D::new(0.0, 0.0)}.
 Notably, methods in Rust are dispatched statically, making them
-eligable for inlining and aggressive optimization; methods are never
+eligible for inlining and aggressive optimization; methods are never
 dispatched dynamically through a function pointer in Rust unless
@@
 
-Finally, Rust has a hygeinic macro system, much more powerful than
+Finally, Rust has a hygienic macro system, much more powerful than
 the C preprocessor. Macro invocations look like function invocations
@@
 Rust's novelty though is in \emph{ownership and borrowing}.
-In Rust, each value is uniquly owned by a single variable, and
+In Rust, each value is uniquely owned by a single variable, and
 assigning that value to another transfers ownership of, and access to,

### 2015-10-24 icse16-servo/rust.tex
minor text and comment tweaks

@@
 instances of enums cannot be indiscriminately cast between
-variants; to access the values in an enum one must employ the \lstinline[language=Rust]{match}
+variants.
+To access the values in an enum one must employ the \lstinline[language=Rust]{match}
 expression to first check the variant, then bind references
@@
 Notably, methods in Rust are dispatched statically, making them
-eligible for inlining and aggressive optimization; methods are never
-dispatched dynamically through a function pointer in Rust unless
-explicitly requested.
+eligible for inlining and aggressive optimization.
+Methods are never dispatched dynamically through a function pointer in Rust
+unless explicitly requested.
 
@@
 where the function name is appended with a bang (\lstinline[language=Rust]{!}).
-They are not functions though, but are expanded in place at compile time,
+They are not functions, but are instead expanded in place at compile time,
 and perform syntactic transformations unavailable to functions.
-The common method of performing console output, \lstinline[language=Rust]~println!(``{}'', foo)~,
+The common method of performing console output, \lstinline[language=Rust]~println!("{}", foo)~,
 is a macro.
@@
 The features of Rust described so far are common to many languages.
-Rust's novelty though is in \emph{ownership and borrowing}.
+Rust's novelty is in \emph{ownership and borrowing}.
 In Rust, each value is uniquely owned by a single variable, and
-assigning that value to another transfers ownership of, and access to,
+assigning that value to another variable transfers ownership of, and access to,
 that value to the receiver. Rust calls this transfer of ownership a \emph{move}.
@@
 
-To make using variables usably ergonomic Rust can create temporary
+To make using variables ergonomic, Rust can create temporary
 references to values with the ampersand (\lstinline[language=Rust]{&})
@@
 
-// Move the value in a to a new location
+// Move the value in `val` to a new location
 let moved_val = val;

### 2015-10-24 icse16-servo/servo.tex
Add conclusion. Fix typos.
Add jdm feedback. Fix urls in footnotes.

@@
 
+Rust also requires that all memory is initialized.
+Failure to initialize memory also has led to crashes in Firefox.\footnote{\url{https://bugzilla.mozilla.org/show_bug.cgi?id=1088731}}
+
 One area for future improvement is related to allocations that are not owned by Rust itself.
@@
 We have implemented wrapper types and compiler plugins that restrict incorrect uses of these foreign values,
-but they are still a source of bugs and one of our largest areas of unsafe code.\footnote{https://blog.mozilla.org/research/2014/08/26/javascript-servos-only-garbage-collector/}
+but they are still a source of bugs and one of our largest areas of unsafe code.\footnote{\url{https://blog.mozilla.org/research/2014/08/26/javascript-servos-only-garbage-collector/}}
 
@@
 In the future, we hope to expose our Rust libraries written for Servo to C, in order to reuse them
-in the Firefox web brower.
-A first potential library is our Rust-based URL parser.\footnote{https://github.com/servo/rust-url}
-While URL parsing seems like a simple and secure task, an audit of the public security bugs\footnote{Firefox public security bugs: http://mzl.la/1GkOiIC} shows that all four of the ones that come from URL parsing are related to either indexing out of
+in the Firefox web browser.
+A first potential library is our Rust-based URL parser.\footnote{\url{https://github.com/servo/rust-url}}
+While URL parsing seems like a simple and secure task, an audit of the public security bugs\footnote{Firefox public security bugs: \url{http://mzl.la/1GkOiIC}} shows that all four of the ones that come from URL parsing are related to either indexing out of
 range of a raw memory buffer or using data that had already been freed --- both of which are

### 2016-01-12 icse16-servo/servo.tex
Clarify the abstraction review point

@@
 that works across all platforms.
-Therefore, much like other modern browsers, Servo contains many of its own specialized implementations of
-library functions that are tuned for the specific cases of web browsers.
+Much like other modern browsers, Servo contains many of its own specialized implementations of
+library functions that are tuned for the specific use cases of web browsers.
+This exhibits the control that Rust provides the programmer -- high level abstractions are available, but should the need arise for a specialized abstraction, it can be written with ease by using lower-level constructs.
 For example, we have special ``small'' vectors that allow instantiation with a default inline size,

### 2016-02-08 icse16-servo/concl.tex

@@
 programming language to develop Servo.
-Based on our experiences to date, we would recommend that either any new systems
-project or new submodule within a larger C or \Cplusplus{} systems project strongly
-consider the use of Rust as an implementation language.
+We cannot yet quantify the amount of additional work that is required by
+the stricter typesystem to author Rust code instead of C or \Cplusplus{}.
+But, we have found that the Rust language is accessible to new developers who
+only have experience in \Cplusplus{}, Python, or JavaScript --- even for writing
+core performance- and memory-sensitive libraries.
+Further, with a relatively small team and short time, we have written a browser
+engine that is increasingly competitive both in performance and functionality
+with other modern browser engines.
+
+The strongest evidence that Rust is not too expensive to write, though, is that
+it can statically verify the absence of memory safety errors.
+Other systems projects --- including all currently shipping browser engines --
+have spent great amounts of time writing and reviewing their C or \Cplusplus{}
+code, written fuzzers, and used internal and external static analysis tools, but
+their security vulnerabilities are still primarily memory safety errors.
+For projects where memory safety errors need to be eliminated, Rust is a
+superior choice.
